---
title: 'Inflation Dashboard'
author: 'Shria Holla, Kelli Marquardt, and Aryan Safi'
date: today
format: html
editor: visual
code-fold: true
execute:
  warning: false
  message: false
toc: true
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  fig.width = 12,   # width in inches
  fig.height = 6,  # height in inches
  echo = TRUE,     # show code by default
  warning = FALSE, # hide warnings
  message = FALSE  # hide messages
)
# read in the population data from ipums
pacman::p_load(here,
               tidyverse,
               data.table,
               importhaver,
               Haver,
               readxl,
               fredr)

# set theme
theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 22, face = 'bold'),
      plot.subtitle = element_text(size = 20),
      legend.title = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.position = 'bottom',
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 18),
      plot.caption = element_text(size = 12),
      axis.line = element_line(color = "black"),
      strip.text = element_text(size = 16),
      axis.ticks = element_line(size = 1.01)
    )
)

```

## Healthcare Services Price Index Summary

This section tracks headline healthcare services price indices from three major U.S. statistical programs. Each table reports the price index level, year-over-year (YoY) change, and month-over-month (MoM) change for the most recent four months and a trailing 12-month average. Because these series are released on different schedules, "Month $t$" reflects each source's own most recent available data.

For a given price index $P_t$, the measures are computed as:

$$
\text{YoY}_t = \left(\frac{P_t}{P_{t-12}} - 1\right) \times 100
$$

$$
\text{MoM}_t = \left(\frac{P_t}{P_{t-1}} - 1\right) \times 100
$$

The four headline series are:

| Source | Haver Code | Description |
|------------------|--------------------------|----------------------------|
| PCE Price Index | `jcsdm@USNA` | Healthcare Services PCE Price Index (SA) |
| CPI | `UMS@CPIDATA` | CPI-U Medical Care Services (SA) |
| PPI Commodity | `P51@PPI` | PPI Healthcare Services, Commodity-Based (NSA) |
| PPI Industry | `RHC@PPIR` | PPI Selected Health Care Industries, Industry-Based (SA) |

```{r summary-helper}
# --------------------------------------------------------------------------
# Helper: build a summary table for a single headline price index series
# --------------------------------------------------------------------------
build_summary_table <- function(series_code, database, label) {
  full_code <- paste0(series_code, "@", database)
  d <- import_haver(series = full_code, eop = FALSE)
  col <- setdiff(names(d), "date")
  names(d)[names(d) == col[1]] <- "pi"

  d <- d |>
    arrange(date) |>
    filter(!is.na(pi)) |>
    mutate(
      yoy = (pi / lag(pi, 12) - 1) * 100,
      mom = (pi / lag(pi, 1) - 1) * 100
    )

  # Most recent month for this series
  latest <- d |> filter(!is.na(yoy)) |> slice_tail(n = 1)
  latest_date <- latest$date

  # Get the 4 most recent months with complete data
  recent <- d |>
    filter(!is.na(yoy), date <= latest_date) |>
    slice_tail(n = 4)

  # 12-month average (t-1 to t-13, excluding the most recent month)
  trailing_12 <- d |>
    filter(!is.na(yoy), date <= latest_date) |>
    slice_tail(n = 13) |>
    slice_head(n = 12)

  avg_pi  <- mean(trailing_12$pi)
  avg_yoy <- mean(trailing_12$yoy)
  avg_mom <- mean(trailing_12$mom, na.rm = TRUE)

  # Build the display columns: t, t-1, t-2, t-3, 12-mo avg
  col_labels <- c(
    format(recent$date[4], "%b %Y"),
    format(recent$date[3], "%b %Y"),
    format(recent$date[2], "%b %Y"),
    format(recent$date[1], "%b %Y"),
    "12-Mo Avg"
  )

  tbl <- data.frame(
    Measure = c("Price Index", "YoY Change (%)", "MoM Change (%)"),
    t0 = c(
      round(recent$pi[4], 2),
      round(recent$yoy[4], 2),
      round(recent$mom[4], 2)
    ),
    t1 = c(
      round(recent$pi[3], 2),
      round(recent$yoy[3], 2),
      round(recent$mom[3], 2)
    ),
    t2 = c(
      round(recent$pi[2], 2),
      round(recent$yoy[2], 2),
      round(recent$mom[2], 2)
    ),
    t3 = c(
      round(recent$pi[1], 2),
      round(recent$yoy[1], 2),
      round(recent$mom[1], 2)
    ),
    avg = c(
      round(avg_pi, 2),
      round(avg_yoy, 2),
      round(avg_mom, 2)
    )
  )

  names(tbl) <- c("Measure", col_labels)

  list(table = tbl, label = paste0(label, " (", series_code, "@", database, ")"))
}
```

### PCE Price Index: Healthcare Services

```{r summary-pce}
pce <- build_summary_table("jcsdm", "USNA", "PCE Healthcare Services")
knitr::kable(pce$table, caption = pce$label, align = c("l", rep("r", 5)))
```

### CPI: Medical Care Services

```{r summary-cpi}
cpi <- build_summary_table("UMS", "CPIDATA", "CPI Medical Care Services")
knitr::kable(cpi$table, caption = cpi$label, align = c("l", rep("r", 5)))
```

### PPI Commodity: Healthcare Services

```{r summary-ppi-commodity}
ppi_comm <- build_summary_table("P51", "PPI", "PPI Healthcare Services (Commodity)")
knitr::kable(ppi_comm$table, caption = ppi_comm$label, align = c("l", rep("r", 5)))
```

### PPI Industry: Selected Health Care Industries

```{r summary-ppi-industry}
ppi_ind <- build_summary_table("RHC", "PPIR", "PPI Selected Health Care Industries")
knitr::kable(ppi_ind$table, caption = ppi_ind$label, align = c("l", rep("r", 5)))
```

## Healthcare Services Contribution to Core PCE Inflation

This section measures how much of Core PCE inflation is attributable to healthcare services. The approach follows Shria Holla's code in `healthcare_graphs.Rmd` (section 5).

Core PCE excludes food and energy and can be decomposed into three mutually exclusive components:

| Component | Haver NE Code | Haver PI Code | Description |
|-----------------|-------------------|-------------------|-----------------|
| Goods ex Food & Energy | `cgxfem@USNA` | `jcgxfem@USNA` | Core goods (excluding food and energy) |
| Services ex Energy & Healthcare | `csxhem@USNA` | `jcsxhem@USNA` | Core services (excluding energy and healthcare) |
| Healthcare Services | `csdm@USNA` | `jcsdm@USNA` | PCE healthcare services |
| **Core PCE (aggregate)** | `cxfem@USNA` | `jcxfem@USNA` | Core PCE total |

For each component $i$, we first compute the **YoY inflation rate** from its price index $P_{i,t}$:

$$
\pi_{i,t} = \left(\frac{P_{i,t}}{P_{i,t-12}} - 1\right) \times 100
$$

Next, we compute each component's **expenditure share** of Core PCE using nominal expenditures ($NE$):

$$
w_{i,t} = \frac{NE_{i,t}}{NE_{core,t}}
$$

Each component's **contribution** to Core PCE inflation (in percentage points) is the product of its share and its inflation rate:

$$
C_{i,t} = w_{i,t} \times \pi_{i,t}
$$

**Estimated Core PCE inflation** is the sum of all component contributions:

$$
\hat{\pi}_{core,t} = C_{goodxfe,t} + C_{servxeh,t} + C_{hcserv,t}
$$

Finally, **healthcare's share of Core PCE inflation** expresses healthcare's contribution as a percentage of the total:

$$
S_{hcserv,t} = \frac{C_{hcserv,t}}{\hat{\pi}_{core,t}} \times 100
$$

```{r core-pce-contribution}
# --------------------------------------------------------------------------
# Step 1: Pull nominal expenditures and price indices from Haver
# --------------------------------------------------------------------------
# Core PCE is decomposed into three components:
#   - goodxfe: Goods excluding Food & Energy
#   - servxeh: Services excluding Energy & Healthcare
#   - hcserv:  Healthcare Services
#
# For each we need nominal expenditures (NE) to compute shares,
# and price indices (PI) to compute inflation rates.

core_data <- import_haver(
  series = c(
    "cxfem@USNA",   # Core PCE NE
    "cgxfem@USNA",  # Goods ex F&E NE
    "csxhem@USNA",  # Services ex E&HC NE
    "csdm@USNA",    # HC Services NE
    "jcxfem@USNA",  # Core PCE PI
    "jcgxfem@USNA", # Goods ex F&E PI
    "jcsxhem@USNA", # Services ex E&HC PI
    "jcsdm@USNA"    # HC Services PI
  ),
  eop = FALSE
) |>
  rename(
    core_ne    = cxfem,
    goodxfe_ne = cgxfem,
    servxeh_ne = csxhem,
    hcserv_ne  = csdm,
    core_pi    = jcxfem,
    goodxfe_pi = jcgxfem,
    servxeh_pi = jcsxhem,
    hcserv_pi  = jcsdm
  ) |>
  arrange(date)

# --------------------------------------------------------------------------
# Step 2: Compute YoY inflation for each component
# --------------------------------------------------------------------------
# YoY inflation = ((PI_t / PI_{t-12}) - 1) * 100
# This gives the percentage change in prices over the past 12 months.

components <- c("core", "goodxfe", "servxeh", "hcserv")

for (comp in components) {
  pi_col   <- paste0(comp, "_pi")
  infl_col <- paste0(comp, "_infl")
  core_data <- core_data |>
    mutate(!!infl_col := ((get(pi_col) / lag(get(pi_col), 12)) - 1) * 100)
}

# --------------------------------------------------------------------------
# Step 3: Compute expenditure shares relative to Core PCE
# --------------------------------------------------------------------------
# Each component's share = its nominal expenditure / Core PCE nominal expenditure.
# These shares reflect how much of Core PCE spending goes to each category.

for (comp in c("goodxfe", "servxeh", "hcserv")) {
  ne_col    <- paste0(comp, "_ne")
  share_col <- paste0(comp, "_expshare")
  core_data <- core_data |>
    mutate(!!share_col := get(ne_col) / core_ne)
}

# --------------------------------------------------------------------------
# Step 4: Compute contributions (share * inflation)
# --------------------------------------------------------------------------
# A component's contribution to aggregate inflation = its expenditure share
# multiplied by its own inflation rate. This tells us how many percentage
# points of Core PCE inflation are driven by each component.

for (comp in c("goodxfe", "servxeh", "hcserv")) {
  infl_col    <- paste0(comp, "_infl")
  share_col   <- paste0(comp, "_expshare")
  contrib_col <- paste0(comp, "_contrib")
  core_data <- core_data |>
    mutate(!!contrib_col := get(share_col) * get(infl_col))
}

# --------------------------------------------------------------------------
# Step 5: Estimate total Core PCE inflation from contributions
# --------------------------------------------------------------------------
# Summing the three contributions gives an estimate of Core PCE inflation.
# This should closely match the directly computed core_infl.

core_data <- core_data |>
  mutate(core_infl_est = goodxfe_contrib + servxeh_contrib + hcserv_contrib)

# --------------------------------------------------------------------------
# Step 6: Healthcare's share of Core PCE inflation
# --------------------------------------------------------------------------
# What percentage of total Core PCE inflation is attributable to healthcare?
# hcserv_contrib_share = (hcserv_contrib / core_infl_est) * 100

core_data <- core_data |>
  mutate(hcserv_contrib_share = (hcserv_contrib / core_infl_est) * 100)
```

### Summary Table

This table displays healthcare services' contribution to Core PCE inflation for the most recent months and a 12-month trailing average. The "Contribution (pp)" row shows how many percentage points of Core PCE inflation come from healthcare services. The "Share of Core PCE Inflation (%)" row shows healthcare's contribution as a percentage of total Core PCE inflation.

```{r core-pce-table}
# --------------------------------------------------------------------------
# Build the summary table for the same time periods as the other tables
# --------------------------------------------------------------------------
contrib_data <- core_data |>
  filter(!is.na(hcserv_contrib_share) & !is.na(core_infl_est)) |>
  select(date, core_infl_est, hcserv_contrib, hcserv_contrib_share)

recent <- contrib_data |> slice_tail(n = 4)
trailing_12 <- contrib_data |> slice_tail(n = 13) |> slice_head(n = 12)

col_labels <- c(
  format(recent$date[4], "%b %Y"),
  format(recent$date[3], "%b %Y"),
  format(recent$date[2], "%b %Y"),
  format(recent$date[1], "%b %Y"),
  "12-Mo Avg"
)

contrib_tbl <- data.frame(
  Measure = c(
    "Core PCE Inflation (%, est.)",
    "HC Services Contribution (pp)",
    "Share of Core PCE Inflation (%)"
  ),
  t0 = c(
    round(recent$core_infl_est[4], 2),
    round(recent$hcserv_contrib[4], 2),
    round(recent$hcserv_contrib_share[4], 2)
  ),
  t1 = c(
    round(recent$core_infl_est[3], 2),
    round(recent$hcserv_contrib[3], 2),
    round(recent$hcserv_contrib_share[3], 2)
  ),
  t2 = c(
    round(recent$core_infl_est[2], 2),
    round(recent$hcserv_contrib[2], 2),
    round(recent$hcserv_contrib_share[2], 2)
  ),
  t3 = c(
    round(recent$core_infl_est[1], 2),
    round(recent$hcserv_contrib[1], 2),
    round(recent$hcserv_contrib_share[1], 2)
  ),
  avg = c(
    round(mean(trailing_12$core_infl_est), 2),
    round(mean(trailing_12$hcserv_contrib), 2),
    round(mean(trailing_12$hcserv_contrib_share), 2)
  )
)

names(contrib_tbl) <- c("Measure", col_labels)

knitr::kable(
  contrib_tbl,
  caption = "Healthcare Services Contribution to Core PCE Inflation (cxfem, cgxfem, csxhem, csdm, jcxfem, jcgxfem, jcsxhem, jcsdm @USNA)",
  align = c("l", rep("r", 5))
)
```

## Medical Care Services Contribution to Core CPI Inflation

This section measures how much of Core CPI inflation is attributable to medical care services, mirroring the PCE contribution analysis above.

Core CPI (All Items Less Food & Energy) includes medical care services as a sub-component. We measure medical care services' contribution using BLS relative importance weights, which reflect each component's expenditure share of the total index.

| Series | Haver Code | Description |
|--------|-----------|-------------|
| Core CPI Price Index | `UIXFDG@CPIDATA` | CPI-U All Items Less Food & Energy (SA) |
| Medical Care Services Price Index | `UMS@CPIDATA` | CPI-U Medical Care Services (SA) |
| Medical Care Services RI Weight | `RPCUMS@CPIDATA` | Relative importance weight, Medical Care Services |
| Core CPI RI Weight | `RUIXFDGM@CPIDATA` | Relative importance weight, All Items Less Food & Energy |

For each price index $P_t$, we compute year-over-year inflation:

$$
\pi_t = \left(\frac{P_t}{P_{t-12}} - 1\right) \times 100
$$

Medical care services' weight within Core CPI is obtained by rescaling the BLS relative importance weights (which are shares of All Items CPI) to shares of Core CPI:

$$
w_{ms,t} = \frac{RI^{ms}_t}{RI^{core}_t}
$$

The contribution (in percentage points) and share of Core CPI inflation are:

$$
C_{ms,t} = w_{ms,t} \times \pi_{ms,t}, \qquad S_{ms,t} = \frac{C_{ms,t}}{\pi_{core,t}} \times 100
$$

```{r core-cpi-contribution}
# --------------------------------------------------------------------------
# Step 1: Pull price indices and relative importance weights from Haver
# --------------------------------------------------------------------------
cpi_core_data <- import_haver(
  series = c(
    "UIXFDG@CPIDATA",   # Core CPI PI (All Items Less Food & Energy, SA)
    "UMS@CPIDATA",       # Medical Care Services PI (SA)
    "RPCUMS@CPIDATA",    # RI weight: Medical Care Services
    "RUIXFDGM@CPIDATA"   # RI weight: All Items Less Food & Energy
  ),
  eop = FALSE
) |>
  rename(
    core_pi    = uixfdg   ,
    ms_pi      = ums,
    ms_ri      = rpcums,
    core_ri    = ruixfdgm
  ) |>
  arrange(date)

# --------------------------------------------------------------------------
# Step 2: Compute YoY inflation for Core CPI and Medical Care Services
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(
    core_infl = ((core_pi / lag(core_pi, 12)) - 1) * 100,
    ms_infl   = ((ms_pi / lag(ms_pi, 12)) - 1) * 100
  )

# --------------------------------------------------------------------------
# Step 3: Rescale RI weights to get Medical Care Services' share of Core CPI
# --------------------------------------------------------------------------
# BLS relative importance weights are shares of All Items CPI.
# Dividing by the Core CPI weight gives the share within Core CPI.
cpi_core_data <- cpi_core_data |>
  mutate(ms_weight = ms_ri / core_ri)

# --------------------------------------------------------------------------
# Step 4: Contribution = weight * component inflation
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(ms_contrib = ms_weight * ms_infl)

# --------------------------------------------------------------------------
# Step 5: Medical Care Services' share of Core CPI inflation
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(ms_contrib_share = (ms_contrib / core_infl) * 100)
```

### Summary Table

This table displays medical care services' contribution to Core CPI inflation for the most recent months and a 12-month trailing average. The "Contribution (pp)" row shows how many percentage points of Core CPI inflation come from medical care services. The "Share of Core CPI Inflation (%)" row shows medical care services' contribution as a percentage of total Core CPI inflation.

```{r core-cpi-table}
# --------------------------------------------------------------------------
# Build the summary table for the same time periods as the other tables
# --------------------------------------------------------------------------
cpi_contrib_data <- cpi_core_data |>
  filter(!is.na(ms_contrib_share) & !is.na(core_infl)) |>
  select(date, core_infl, ms_contrib, ms_contrib_share)

recent_cpi <- cpi_contrib_data |> slice_tail(n = 4)
trailing_12_cpi <- cpi_contrib_data |> slice_tail(n = 13) |> slice_head(n = 12)

col_labels_cpi <- c(
  format(recent_cpi$date[4], "%b %Y"),
  format(recent_cpi$date[3], "%b %Y"),
  format(recent_cpi$date[2], "%b %Y"),
  format(recent_cpi$date[1], "%b %Y"),
  "12-Mo Avg"
)

cpi_contrib_tbl <- data.frame(
  Measure = c(
    "Core CPI Inflation (%)",
    "MC Services Contribution (pp)",
    "Share of Core CPI Inflation (%)"
  ),
  t0 = c(
    round(recent_cpi$core_infl[4], 2),
    round(recent_cpi$ms_contrib[4], 2),
    round(recent_cpi$ms_contrib_share[4], 2)
  ),
  t1 = c(
    round(recent_cpi$core_infl[3], 2),
    round(recent_cpi$ms_contrib[3], 2),
    round(recent_cpi$ms_contrib_share[3], 2)
  ),
  t2 = c(
    round(recent_cpi$core_infl[2], 2),
    round(recent_cpi$ms_contrib[2], 2),
    round(recent_cpi$ms_contrib_share[2], 2)
  ),
  t3 = c(
    round(recent_cpi$core_infl[1], 2),
    round(recent_cpi$ms_contrib[1], 2),
    round(recent_cpi$ms_contrib_share[1], 2)
  ),
  avg = c(
    round(mean(trailing_12_cpi$core_infl), 2),
    round(mean(trailing_12_cpi$ms_contrib), 2),
    round(mean(trailing_12_cpi$ms_contrib_share), 2)
  )
)

names(cpi_contrib_tbl) <- c("Measure", col_labels_cpi)

knitr::kable(
  cpi_contrib_tbl,
  caption = "Medical Care Services Contribution to Core CPI Inflation (UIXFDG, UMS, RPCUMS, RUIXFDGM @CPIDATA)",
  align = c("l", rep("r", 5))
)
```

## Data Staleness Monitor

This section checks the last available date for every Haver series used across the project's PCE, PPI, and CPI code. Series whose end date matches the most recent available month are highlighted in green ("Current"); those 1 month behind are in yellow ("Behind"); those 2–5 months behind are in orange ("Delayed"); and those 6 or more months behind (or missing entirely) are in red ("Stale").

```{r staleness-monitor}
# --------------------------------------------------------------------------
# Define all Haver series used across the project, grouped by source
# --------------------------------------------------------------------------
series_registry <- tribble(
  ~series_code,   ~database,   ~source_group, ~description,

  # --- PCE Nominal Expenditures (USNA) --- (SA, Mil$). NE = nominal expenditures
  "cm",           "USNA",      "PCE",         "Overall PCE NE",
  "cxfem",        "USNA",      "PCE",         "Core PCE NE",
  "cgxfem",       "USNA",      "PCE",         "Goods ex F&E NE",
  "csm",          "USNA",      "PCE",         "Services NE",
  "csxhem",       "USNA",      "PCE",         "Services ex E&HC NE",
  "csdm",         "USNA",      "PCE",         "Healthcare Services NE",
  "csdom",        "USNA",      "PCE",         "Healthcare Services: Outpatient Services NE",
  "csmpm",        "USNA",      "PCE",         "Healthcare Services: Physician Services NE",
  "csmdm",        "USNA",      "PCE",         "Healthcare Services: Dental Services NE",
  "csmom",        "USNA",      "PCE",         "Healthcare Services: Paramedical Services NE",
  "csmnm",        "USNA",      "PCE",         "Healthcare Services: Paramedical: Home Health Care NE",
  "csmolm",       "USNA",      "PCE",         "Healthcare Services: Paramedical: Medical Laboratories NE",
  "csmoom",       "USNA",      "PCE",         "Healthcare Services: Paramedical: Other Professional Medical Services NE",
  "csmhpm",       "USNA",      "PCE",         "Healthcare Services: Hospital Services NE",
  "csmhnm",       "USNA",      "PCE",         "Healthcare Services: Nursing Homes NE",
  "ctgm",         "USNA",      "PCE",         "Total Goods NE",
  "csmoa",        "USNA",      "PCE",         "Healthcare Services: Paramedical Services",

  # --- PCE Price Indices (USNA) --- (seasonally adjusted= SA)
  "jcm",          "USNA",      "PCE",         "Overall PCE PI",
  "jcxfem",       "USNA",      "PCE",         "Core PCE PI (ex F&E)",
  "jcgxfem",      "USNA",      "PCE",         "Goods ex F&E PCE PI",
  "jcsm",         "USNA",      "PCE",         "Services PCE PI",
  "jcsxhem",      "USNA",      "PCE",         "Services ex E&HC PCE PI",
  "jcsdm",        "USNA",      "PCE",         "Healthcare Services PCE PI",
  "jcsdom",       "USNA",      "PCE",         "Outpatient Services PCE PI",
  "jcsmpm",       "USNA",      "PCE",         "Physician Services PCE PI",
  "jcsmdm",       "USNA",      "PCE",         "Dental Services PCE PI",
  "jcsmom",       "USNA",      "PCE",         "Paramedical Services PCE PI",
  "jcsmnm",       "USNA",      "PCE",         "Home & Nursing PCE PI",
  "jcsmhpm",      "USNA",      "PCE",         "Hospital Services PCE PI",
  "jcsmhnm",      "USNA",      "PCE",         "Nursing Homes PCE PI",
  "jctgm",        "USNA",      "PCE",         "Total Goods PCE PI",

  # --- Core PCE (USECON) --- (SA)
  "jcxfebm",      "USECON",    "PCE",         "Core PCE PI (USECON)",

  # --- PPI Industry-Based (PPIR) --- (appear to be SA)
  "RHC",          "PPIR",      "PPI",         "PPI Selected Health Care Industries",
  "R6211",        "PPIR",      "PPI",         "PPI Offices of Physicians",
  "R6212",        "PPIR",      "PPI",         "PPI Offices of Dentists",
  "R6215",        "PPIR",      "PPI",         "PPI Medical & Diagnostic Labs",
  "R6216",        "PPIR",      "PPI",         "PPI Home Health",
  "R621991",      "PPIR",      "PPI",         "PPI Blood and Organ Banks",
  "R622",         "PPIR",      "PPI",         "PPI Hospitals",
  "R6231",        "PPIR",      "PPI",         "PPI Nursing Facilities",
  "R62321",       "PPIR",      "PPI",         "PPI Residential Developmental Disability Homes",

  # special index (table 11) (NSA)
  "R62211A2",     "PPIR",      "PPI",         "PPI Hospitals - Medicare",
  "R62211A4",     "PPIR",      "PPI",         "PPI Hospitals - Medicaid",
  "R62211A6",     "PPIR",      "PPI",         "PPI Hospitals - Private/Other",

  # --- PPI Commodity-Based (PPI) --- (only pulling NSA because it's the most available)
  "P51",          "PPI",       "PPI",         "PPI Healthcare Services (Commodity)",
  "P511",         "PPI",       "PPI",         "PPI Healthcare Services: Outpatient (Commodity)",
  "P511104B",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Medicare (Commodity)",
  "P511104C",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Medicaid (Commodity)",
  "P511104D",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Private/All Other (Commodity)",

  "P512",         "PPI",       "PPI",         "PPI Healthcare Services: Inpatient (Commodity)",
  "P512101B",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Medicare (Commodity)",
  "P512101C",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Medicaid (Commodity)",
  "P512101D",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Private/All Other (Commodity)",

  # --- CPI (CPIDATA) ---
  "UM",           "CPIDATA",   "CPI",         "CPI-U Medical Care (SA)",
  "UMC",          "CPIDATA",   "CPI",         "CPI-U Medical Care Commodities (SA)",
  "UMG",          "CPIDATA",   "CPI",         "CPI-U Medicinal Drugs (SA)",
  "UMP",          "CPIDATA",   "CPI",         "CPI-U Prescription Drugs (SA)",
  "UMGNA",        "CPIDATA",   "CPI",         "CPI-U Nonprescription Drugs (SA)",
  "UMS",          "CPIDATA",   "CPI",         "CPI-U Medicare Services (SA)",
  "UMSPP",        "CPIDATA",   "CPI",         "CPI-U Physicians' Services (SA)",
  "UMSPT",        "CPIDATA",   "CPI",         "CPI-U Dental Services (SA)",
  "UMSPE",        "CPIDATA",   "CPI",         "CPI-U Eyeglasses & Eyecare (SA)",
  "UMSPS",        "CPIDATA",   "CPI",         "CPI-U Services by Other Medical Professionals (SA)",
  "UMSH",         "CPIDATA",   "CPI",         "CPI-U Hospital and Related Services (SA)",
  "UMSHS",        "CPIDATA",   "CPI",         "CPI-U Hospital Services (SA)",
  "UMSHSI",       "CPIDATA",   "CPI",         "CPI-U Inpatient Hospital Services (SA)",
  "UMSHSO",       "CPIDATA",   "CPI",         "CPI-U Outpatient Hospital Services (SA)",
  "UMSNS",        "CPIDATA",   "CPI",         "CPI-U Nursing Homes and Adult Daycare (SA)",
  # Not seasonally adjusted
  "UMN",          "CPIDATA",   "CPI",         "CPI-U Medical Care (NSA)",
  "UMCN",         "CPIDATA",   "CPI",         "CPI-U Medical Care Commodities (NSA)",
  "UMGN",         "CPIDATA",   "CPI",         "CPI-U Medicinal Drugs (NSA)",
  "UMPN",         "CPIDATA",   "CPI",         "CPI-U Prescription Drugs (NSA)",
  "UMGNN",        "CPIDATA",   "CPI",         "CPI-U Nonprescription Drugs (NSA)",
  "UMQN",         "CPIDATA",   "CPI",         "CPI-U Medical Equipment and Supplies (NSA)",
  "UMSN",         "CPIDATA",   "CPI",         "CPI-U Medical Care Services (NSA)",
  "UMSPN",        "CPIDATA",   "CPI",         "CPI-U Professional Medical Care Services (NSA)",
  "UMSPPN",       "CPIDATA",   "CPI",         "CPI-U Physicians' Services (NSA)",
  "UMSPTN",       "CPIDATA",   "CPI",         "CPI-U Dental Services (NSA)",
  "UMSPEN",       "CPIDATA",   "CPI",         "CPI-U Eyeglasses and Eye Care (NSA)",
  "UMSPSN",       "CPIDATA",   "CPI",         "CPI-U Services by Other Medical Professionals (NSA)",
  "UMSHN",        "CPIDATA",   "CPI",         "CPI-U Hospital and Related Services (NSA)",
  "UMSHSN",       "CPIDATA",   "CPI",         "CPI-U Hospital Services (NSA)",
  "UMSHSIN",      "CPIDATA",   "CPI",         "CPI-U Inpatient Hospital Services (NSA)",
  "UMSHSON",      "CPIDATA",   "CPI",         "CPI-U Outpatient Hospital Services (NSA)",
  "UMSNSN",       "CPIDATA",   "CPI",         "CPI-U Nursing Homes and Adult Daycare (NSA)",
  "UOEECN",       "CPIDATA",   "CPI",         "CPI-U Home Health Care",
  "UMSIN",        "CPIDATA",   "CPI",         "CPI-U Health Insurance",
  "UIXFDG",       "CPIDATA",   "CPI",         "CPI-U All Items Less Food & Energy (SA)",
  "RPCUMS",       "CPIDATA",   "CPI",         "CPI-U RI Weight: Medical Care Services",
  "RUIXFDGM",     "CPIDATA",   "CPI",         "CPI-U RI Weight: All Items Less Food & Energy"
)

# --------------------------------------------------------------------------
# Pull each series and extract the last non-NA date
# --------------------------------------------------------------------------
get_end_date <- function(code, db) {
  full_code <- paste0(code, "@", db)
  tryCatch({
    d <- import_haver(series = full_code, eop = FALSE) |>
      na.omit()
    col <- setdiff(names(d), "date")
    d <- d |> filter(!is.na(.data[[col[1]]]))
    if (nrow(d) == 0) return(as.Date(NA))
    max(d$date)
  }, error = function(e) as.Date(NA))
}

staleness <- series_registry |>
  rowwise() |>
  mutate(end_date = get_end_date(series_code, database)) |>
  ungroup()

# --------------------------------------------------------------------------
# Determine the most recent end date and flag stale series
# --------------------------------------------------------------------------
most_recent <- max(staleness$end_date, na.rm = TRUE)

staleness <- staleness |>
  mutate(
    months_behind = as.numeric(
      (year(most_recent) - year(end_date)) * 12 +
      (month(most_recent) - month(end_date))
    ),
    status = case_when(
      is.na(end_date)       ~ "Stale",
      months_behind == 0    ~ "Current",
      months_behind == 1    ~ "Behind",
      months_behind <= 5    ~ "Delayed",
      TRUE                  ~ "Stale"
    )
  )

# --------------------------------------------------------------------------
# Single color-coded table for all series
# --------------------------------------------------------------------------
staleness_display <- staleness |>
  select(source_group, series_code, database, description, end_date, status) |>
  arrange(status, source_group, series_code)

DT::datatable(
  staleness_display,
  caption = paste0(
    "Series Staleness Monitor (most recent: ",
    format(most_recent, "%B %Y"), ")"
  ),
  colnames = c("Source", "Code", "Database", "Description", "End Date", "Status"),
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    scrollY = T,
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, -1), c('10', '25', '50', 'All'))
  )
) |>
  DT::formatStyle(
    "status",
    target = "row",
    backgroundColor = DT::styleEqual(
      c("Current", "Behind", "Delayed", "Stale"),
      c("#d4edda", "#fff3cd", "#ffd6a5", "#f8d7da")
    )
  )
```
