---
title: 'Inflation Dashboard'
author: 'Shria Holla, Kelli Marquardt, and Aryan Safi'
date: today
format: html
editor: visual
code-fold: true
execute:
  warning: false
  message: false
toc: true
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  fig.width = 12,   # width in inches
  fig.height = 6,  # height in inches
  echo = TRUE,     # show code by default
  warning = FALSE, # hide warnings
  message = FALSE  # hide messages
)
# read in the population data from ipums
pacman::p_load(here,
               tidyverse,
               data.table,
               importhaver,
               Haver,
               readxl,
               fredr,
               plotly)

# set theme
theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 22, face = 'bold'),
      plot.subtitle = element_text(size = 20),
      legend.title = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.position = 'bottom',
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 18),
      plot.caption = element_text(size = 12),
      axis.line = element_line(color = "black"),
      strip.text = element_text(size = 16),
      axis.ticks = element_line(size = 1.01)
    )
)

```

## Healthcare Services Price Index Summary

This section tracks headline healthcare services price indices from three major U.S. statistical programs. Each table reports the price index level, year-over-year (YoY) change, and month-over-month (MoM) change for the most recent four months and a trailing 12-month average. Because these series are released on different schedules, "Month $t$" reflects each source's own most recent available data.

For a given price index $P_t$, the measures are computed as:

$$\text{YoY}_t = \left(\frac{P_t}{P_{t-12}} - 1\right) \times 100$$

$$\text{MoM}_t = \left(\frac{P_t}{P_{t-1}} - 1\right) \times 100$$

The four headline series are:

| Source | Haver Code | Description |
|------------------------|------------------------|------------------------|
| PCE Price Index | `jcsdm@USNA` | Healthcare Services PCE Price Index (SA) |
| CPI | `UMS@CPIDATA` | CPI-U Medical Care Services (SA) |
| PPI Commodity | `P51@PPI` | PPI Healthcare Services, Commodity-Based (NSA) |
| PPI Industry | `RHC@PPIR` | PPI Selected Health Care Industries, Industry-Based (SA) |

```{r summary-helper}
# --------------------------------------------------------------------------
# Helper: build a summary table for a single headline price index series
# --------------------------------------------------------------------------
build_summary_table <- function(series_code, database, label) {
  full_code <- paste0(series_code, "@", database)
  d <- import_haver(series = full_code, eop = FALSE)
  col <- setdiff(names(d), "date")
  names(d)[names(d) == col[1]] <- "pi"

  d <- d |>
    arrange(date) |>
    filter(!is.na(pi)) |>
    mutate(
      yoy = (pi / lag(pi, 12) - 1) * 100,
      mom = (pi / lag(pi, 1) - 1) * 100
    )

  # Most recent month for this series
  latest <- d |> filter(!is.na(yoy)) |> slice_tail(n = 1)
  latest_date <- latest$date

  # Get the 4 most recent months with complete data
  recent <- d |>
    filter(!is.na(yoy), date <= latest_date) |>
    slice_tail(n = 4)

  # 12-month average (t-1 to t-13, excluding the most recent month)
  trailing_12 <- d |>
    filter(!is.na(yoy), date <= latest_date) |>
    slice_tail(n = 13) |>
    slice_head(n = 12)

  avg_pi  <- mean(trailing_12$pi)
  avg_yoy <- mean(trailing_12$yoy)
  avg_mom <- mean(trailing_12$mom, na.rm = TRUE)

  # Build the display columns: t, t-1, t-2, t-3, 12-mo avg
  col_labels <- c(
    format(recent$date[4], "%b %Y"),
    format(recent$date[3], "%b %Y"),
    format(recent$date[2], "%b %Y"),
    format(recent$date[1], "%b %Y"),
    "12-Mo Avg"
  )

  tbl <- data.frame(
    Measure = c("Price Index", "YoY Change (%)", "MoM Change (%)"),
    t0 = c(
      round(recent$pi[4], 2),
      round(recent$yoy[4], 2),
      round(recent$mom[4], 2)
    ),
    t1 = c(
      round(recent$pi[3], 2),
      round(recent$yoy[3], 2),
      round(recent$mom[3], 2)
    ),
    t2 = c(
      round(recent$pi[2], 2),
      round(recent$yoy[2], 2),
      round(recent$mom[2], 2)
    ),
    t3 = c(
      round(recent$pi[1], 2),
      round(recent$yoy[1], 2),
      round(recent$mom[1], 2)
    ),
    avg = c(
      round(avg_pi, 2),
      round(avg_yoy, 2),
      round(avg_mom, 2)
    )
  )

  names(tbl) <- c("Measure", col_labels)

  list(table = tbl, label = paste0(label, " (", series_code, "@", database, ")"))
}

# --------------------------------------------------------------------------
# Helper: build three plotly plots (level, YoY, MoM) for a headline series,
# restricted to the most recent 24 months and overlaying the sample mean.
# --------------------------------------------------------------------------
build_summary_plots <- function(series_code, database, label) {
  full_code <- paste0(series_code, "@", database)
  d <- import_haver(series = full_code, eop = FALSE)
  col <- setdiff(names(d), "date")
  names(d)[names(d) == col[1]] <- "pi"

  d <- d |>
    arrange(date) |>
    filter(!is.na(pi)) |>
    mutate(
      yoy = (pi / lag(pi, 12) - 1) * 100,
      mom = (pi / lag(pi, 1) - 1) * 100
    ) |>
    slice_tail(n = 24)

  make_plot <- function(y_var, y_label, plot_title) {
    sample_mean <- mean(d[[y_var]], na.rm = TRUE)
    mean_label  <- paste0("Sample Mean (", round(sample_mean, 2), ")")

    plotly::plot_ly(d, x = ~date) |>
      plotly::add_lines(
        y = ~.data[[y_var]],
        name = y_label,
        line = list(color = "#1f77b4", width = 2)
      ) |>
      plotly::add_lines(
        y = rep(sample_mean, nrow(d)),
        name = mean_label,
        line = list(color = "red", dash = "dash", width = 1.5)
      ) |>
      plotly::layout(
        title = plot_title,
        xaxis = list(title = ""),
        yaxis = list(title = y_label),
        hovermode = "x unified",
        legend = list(orientation = "h", x = 0, y = -0.15)
      )
  }

  list(
    level = make_plot("pi",  "Index Level", paste0(label, ": Price Index Level (Last 24 Months)")),
    yoy   = make_plot("yoy", "YoY (%)",     paste0(label, ": YoY % Change (Last 24 Months)")),
    mom   = make_plot("mom", "MoM (%)",     paste0(label, ": MoM % Change (Last 24 Months)"))
  )
}
```

### PCE Price Index: Healthcare Services

::: panel-tabset
#### Summary Table

```{r summary-pce}
pce <- build_summary_table("jcsdm", "USNA", "PCE Healthcare Services")
knitr::kable(pce$table, caption = pce$label, align = c("l", rep("r", 5)))
```

#### Price Index Level

```{r summary-pce-level}
pce_plots <- build_summary_plots("jcsdm", "USNA", "PCE Healthcare Services")
pce_plots$level
```

#### YoY

```{r summary-pce-yoy}
pce_plots$yoy
```

#### MoM

```{r summary-pce-mom}
pce_plots$mom
```
:::

### CPI: Medical Care Services

::: panel-tabset
#### Summary Table

```{r summary-cpi}
cpi <- build_summary_table("UMS", "CPIDATA", "CPI Medical Care Services")
knitr::kable(cpi$table, caption = cpi$label, align = c("l", rep("r", 5)))
```

#### Price Index Level

```{r summary-cpi-level}
cpi_plots <- build_summary_plots("UMS", "CPIDATA", "CPI Medical Care Services")
cpi_plots$level
```

#### YoY

```{r summary-cpi-yoy}
cpi_plots$yoy
```

#### MoM

```{r summary-cpi-mom}
cpi_plots$mom
```
:::

### PPI Commodity: Healthcare Services

::: panel-tabset
#### Summary Table

```{r summary-ppi-commodity}
ppi_comm <- build_summary_table("P51", "PPI", "PPI Healthcare Services (Commodity)")
knitr::kable(ppi_comm$table, caption = ppi_comm$label, align = c("l", rep("r", 5)))
```

#### Price Index Level

```{r summary-ppi-commodity-level}
ppi_comm_plots <- build_summary_plots("P51", "PPI", "PPI Healthcare Services (Commodity)")
ppi_comm_plots$level
```

#### YoY

```{r summary-ppi-commodity-yoy}
ppi_comm_plots$yoy
```

#### MoM

```{r summary-ppi-commodity-mom}
ppi_comm_plots$mom
```
:::

### PPI Industry: Selected Health Care Industries

::: panel-tabset
#### Summary Table

```{r summary-ppi-industry}
ppi_ind <- build_summary_table("RHC", "PPIR", "PPI Selected Health Care Industries")
knitr::kable(ppi_ind$table, caption = ppi_ind$label, align = c("l", rep("r", 5)))
```

#### Price Index Level

```{r summary-ppi-industry-level}
ppi_ind_plots <- build_summary_plots("RHC", "PPIR", "PPI Selected Health Care Industries")
ppi_ind_plots$level
```

#### YoY

```{r summary-ppi-industry-yoy}
ppi_ind_plots$yoy
```

#### MoM

```{r summary-ppi-industry-mom}
ppi_ind_plots$mom
```
:::

## Healthcare Services Contribution to Core PCE Inflation

This section measures how much of Core PCE inflation is attributable to healthcare services. The approach follows Shria Holla's code in `healthcare_graphs.Rmd` (section 5).

Core PCE excludes food and energy and can be decomposed into three mutually exclusive components:

| Component | Haver NE Code | Haver PI Code | Description |
|------------------|------------------|------------------|------------------|
| Goods ex Food & Energy | `cgxfem@USNA` | `jcgxfem@USNA` | Core goods (excluding food and energy) |
| Services ex Energy & Healthcare | `csxhem@USNA` | `jcsxhem@USNA` | Core services (excluding energy and healthcare) |
| Healthcare Services | `csdm@USNA` | `jcsdm@USNA` | PCE healthcare services |
| **Core PCE (aggregate)** | `cxfem@USNA` | `jcxfem@USNA` | Core PCE total |

For each component $i$, we first compute the **YoY inflation rate** from its price index $P_{i,t}$:

$$\pi_{i,t} = \left(\frac{P_{i,t}}{P_{i,t-12}} - 1\right) \times 100$$

Next, we compute each component's **expenditure share** of Core PCE using nominal expenditures ($NE$):

$$w_{i,t} = \frac{NE_{i,t}}{NE_{core,t}}$$

Each component's **contribution** to Core PCE inflation (in percentage points) is the product of its share and its inflation rate:

$$C_{i,t} = w_{i,t} \times \pi_{i,t}$$

**Estimated Core PCE inflation** is the sum of all component contributions:

$$\hat{\pi}_{core,t} = C_{goodxfe,t} + C_{servxeh,t} + C_{hcserv,t}$$

Finally, **healthcare's share of Core PCE inflation** expresses healthcare's contribution as a percentage of the total:

$$S_{hcserv,t} = \frac{C_{hcserv,t}}{\hat{\pi}_{core,t}} \times 100$$

```{r core-pce-contribution}
# --------------------------------------------------------------------------
# Step 1: Pull nominal expenditures and price indices from Haver
# --------------------------------------------------------------------------
# Core PCE is decomposed into three components:
#   - goodxfe: Goods excluding Food & Energy
#   - servxeh: Services excluding Energy & Healthcare
#   - hcserv:  Healthcare Services
#
# For each we need nominal expenditures (NE) to compute shares,
# and price indices (PI) to compute inflation rates.

core_data <- import_haver(
  series = c(
    "cxfem@USNA",   # Core PCE NE
    "cgxfem@USNA",  # Goods ex F&E NE
    "csxhem@USNA",  # Services ex E&HC NE
    "csdm@USNA",    # HC Services NE
    "jcxfem@USNA",  # Core PCE PI
    "jcgxfem@USNA", # Goods ex F&E PI
    "jcsxhem@USNA", # Services ex E&HC PI
    "jcsdm@USNA"    # HC Services PI
  ),
  eop = FALSE
) |>
  rename(
    core_ne    = cxfem,
    goodxfe_ne = cgxfem,
    servxeh_ne = csxhem,
    hcserv_ne  = csdm,
    core_pi    = jcxfem,
    goodxfe_pi = jcgxfem,
    servxeh_pi = jcsxhem,
    hcserv_pi  = jcsdm
  ) |>
  arrange(date)

# --------------------------------------------------------------------------
# Step 2: Compute YoY inflation for each component
# --------------------------------------------------------------------------
# YoY inflation = ((PI_t / PI_{t-12}) - 1) * 100
# This gives the percentage change in prices over the past 12 months.

components <- c("core", "goodxfe", "servxeh", "hcserv")

for (comp in components) {
  pi_col   <- paste0(comp, "_pi")
  infl_col <- paste0(comp, "_infl")
  core_data <- core_data |>
    mutate(!!infl_col := ((get(pi_col) / lag(get(pi_col), 12)) - 1) * 100)
}

# --------------------------------------------------------------------------
# Step 3: Compute expenditure shares relative to Core PCE
# --------------------------------------------------------------------------
# Each component's share = its nominal expenditure / Core PCE nominal expenditure.
# These shares reflect how much of Core PCE spending goes to each category.

for (comp in c("goodxfe", "servxeh", "hcserv")) {
  ne_col    <- paste0(comp, "_ne")
  share_col <- paste0(comp, "_expshare")
  core_data <- core_data |>
    mutate(!!share_col := get(ne_col) / core_ne)
}

# --------------------------------------------------------------------------
# Step 4: Compute contributions (share * inflation)
# --------------------------------------------------------------------------
# A component's contribution to aggregate inflation = its expenditure share
# multiplied by its own inflation rate. This tells us how many percentage
# points of Core PCE inflation are driven by each component.

for (comp in c("goodxfe", "servxeh", "hcserv")) {
  infl_col    <- paste0(comp, "_infl")
  share_col   <- paste0(comp, "_expshare")
  contrib_col <- paste0(comp, "_contrib")
  core_data <- core_data |>
    mutate(!!contrib_col := get(share_col) * get(infl_col))
}

# --------------------------------------------------------------------------
# Step 5: Estimate total Core PCE inflation from contributions
# --------------------------------------------------------------------------
# Summing the three contributions gives an estimate of Core PCE inflation.
# This should closely match the directly computed core_infl.

core_data <- core_data |>
  mutate(core_infl_est = goodxfe_contrib + servxeh_contrib + hcserv_contrib)

# --------------------------------------------------------------------------
# Step 6: Healthcare's share of Core PCE inflation
# --------------------------------------------------------------------------
# What percentage of total Core PCE inflation is attributable to healthcare?
# hcserv_contrib_share = (hcserv_contrib / core_infl_est) * 100

core_data <- core_data |>
  mutate(hcserv_contrib_share = (hcserv_contrib / core_infl_est) * 100)
```

### Summary Table

This table displays healthcare services' contribution to Core PCE inflation for the most recent months and a 12-month trailing average. The "Contribution (pp)" row shows how many percentage points of Core PCE inflation come from healthcare services. The "Share of Core PCE Inflation (%)" row shows healthcare's contribution as a percentage of total Core PCE inflation.

```{r core-pce-table}
# --------------------------------------------------------------------------
# Build the summary table for the same time periods as the other tables
# --------------------------------------------------------------------------
contrib_data <- core_data |>
  filter(!is.na(hcserv_contrib_share) & !is.na(core_infl_est)) |>
  select(date, core_infl_est, hcserv_contrib, hcserv_contrib_share)

recent <- contrib_data |> slice_tail(n = 4)
trailing_12 <- contrib_data |> slice_tail(n = 13) |> slice_head(n = 12)

col_labels <- c(
  format(recent$date[4], "%b %Y"),
  format(recent$date[3], "%b %Y"),
  format(recent$date[2], "%b %Y"),
  format(recent$date[1], "%b %Y"),
  "12-Mo Avg"
)

contrib_tbl <- data.frame(
  Measure = c(
    "Core PCE Inflation (%, est.)",
    "HC Services Contribution (pp)",
    "Share of Core PCE Inflation (%)"
  ),
  t0 = c(
    round(recent$core_infl_est[4], 2),
    round(recent$hcserv_contrib[4], 2),
    round(recent$hcserv_contrib_share[4], 2)
  ),
  t1 = c(
    round(recent$core_infl_est[3], 2),
    round(recent$hcserv_contrib[3], 2),
    round(recent$hcserv_contrib_share[3], 2)
  ),
  t2 = c(
    round(recent$core_infl_est[2], 2),
    round(recent$hcserv_contrib[2], 2),
    round(recent$hcserv_contrib_share[2], 2)
  ),
  t3 = c(
    round(recent$core_infl_est[1], 2),
    round(recent$hcserv_contrib[1], 2),
    round(recent$hcserv_contrib_share[1], 2)
  ),
  avg = c(
    round(mean(trailing_12$core_infl_est), 2),
    round(mean(trailing_12$hcserv_contrib), 2),
    round(mean(trailing_12$hcserv_contrib_share), 2)
  )
)

names(contrib_tbl) <- c("Measure", col_labels)

knitr::kable(
  contrib_tbl,
  caption = "Healthcare Services Contribution to Core PCE Inflation (cxfem, cgxfem, csxhem, csdm, jcxfem, jcgxfem, jcsxhem, jcsdm @USNA)",
  align = c("l", rep("r", 5))
)
```

## Medical Care Services Contribution to Core CPI Inflation

This section measures how much of Core CPI inflation is attributable to medical care services, mirroring the PCE contribution analysis above.

Core CPI (All Items Less Food & Energy) includes medical care services as a sub-component. We measure medical care services' contribution using BLS relative importance weights, which reflect each component's expenditure share of the total index.

| Series | Haver Code | Description |
|------------------------|------------------------|------------------------|
| Core CPI Price Index | `UIXFDG@CPIDATA` | CPI-U All Items Less Food & Energy (SA) |
| Medical Care Services Price Index | `UMS@CPIDATA` | CPI-U Medical Care Services (SA) |
| Medical Care Services RI Weight | `RPCUMS@CPIDATA` | Relative importance weight, Medical Care Services |
| Core CPI RI Weight | `RUIXFDGM@CPIDATA` | Relative importance weight, All Items Less Food & Energy |

For each price index $P_t$, we compute year-over-year inflation:

$$\pi_t = \left(\frac{P_t}{P_{t-12}} - 1\right) \times 100$$

Medical care services' weight within Core CPI is obtained by rescaling the BLS relative importance weights (which are shares of All Items CPI) to shares of Core CPI:

$$w_{ms,t} = \frac{RI^{ms}_t}{RI^{core}_t}$$

The contribution (in percentage points) and share of Core CPI inflation are:

$$C_{ms,t} = w_{ms,t} \times \pi_{ms,t}, \qquad S_{ms,t} = \frac{C_{ms,t}}{\pi_{core,t}} \times 100$$

```{r core-cpi-contribution}
# --------------------------------------------------------------------------
# Step 1: Pull price indices and relative importance weights from Haver
# --------------------------------------------------------------------------
cpi_core_data <- import_haver(
  series = c(
    "UIXFDG@CPIDATA",   # Core CPI PI (All Items Less Food & Energy, SA)
    "UMS@CPIDATA",       # Medical Care Services PI (SA)
    "RPCUMS@CPIDATA",    # RI weight: Medical Care Services
    "RUIXFDGM@CPIDATA"   # RI weight: All Items Less Food & Energy
  ),
  eop = FALSE
) |>
  rename(
    core_pi    = uixfdg   ,
    ms_pi      = ums,
    ms_ri      = rpcums,
    core_ri    = ruixfdgm
  ) |>
  arrange(date)

# --------------------------------------------------------------------------
# Step 2: Compute YoY inflation for Core CPI and Medical Care Services
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(
    core_infl = ((core_pi / lag(core_pi, 12)) - 1) * 100,
    ms_infl   = ((ms_pi / lag(ms_pi, 12)) - 1) * 100
  )

# --------------------------------------------------------------------------
# Step 3: Rescale RI weights to get Medical Care Services' share of Core CPI
# --------------------------------------------------------------------------
# BLS relative importance weights are shares of All Items CPI.
# Dividing by the Core CPI weight gives the share within Core CPI.
cpi_core_data <- cpi_core_data |>
  mutate(ms_weight = ms_ri / core_ri)

# --------------------------------------------------------------------------
# Step 4: Contribution = weight * component inflation
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(ms_contrib = ms_weight * ms_infl)

# --------------------------------------------------------------------------
# Step 5: Medical Care Services' share of Core CPI inflation
# --------------------------------------------------------------------------
cpi_core_data <- cpi_core_data |>
  mutate(ms_contrib_share = (ms_contrib / core_infl) * 100)
```

### Summary Table

This table displays medical care services' contribution to Core CPI inflation for the most recent months and a 12-month trailing average. The "Contribution (pp)" row shows how many percentage points of Core CPI inflation come from medical care services. The "Share of Core CPI Inflation (%)" row shows medical care services' contribution as a percentage of total Core CPI inflation.

```{r core-cpi-table}
# --------------------------------------------------------------------------
# Build the summary table for the same time periods as the other tables
# --------------------------------------------------------------------------
cpi_contrib_data <- cpi_core_data |>
  filter(!is.na(ms_contrib_share) & !is.na(core_infl)) |>
  select(date, core_infl, ms_contrib, ms_contrib_share, ms_ri, ms_weight)

recent_cpi <- cpi_contrib_data |> slice_tail(n = 4)
trailing_12_cpi <- cpi_contrib_data |> slice_tail(n = 13) |> slice_head(n = 12)

col_labels_cpi <- c(
  format(recent_cpi$date[4], "%b %Y"),
  format(recent_cpi$date[3], "%b %Y"),
  format(recent_cpi$date[2], "%b %Y"),
  format(recent_cpi$date[1], "%b %Y"),
  "12-Mo Avg"
)

cpi_contrib_tbl <- data.frame(
  Measure = c(
    "Core CPI Inflation (%)",
    "MC Services Contribution (pp)",
    "Share of Core CPI Inflation (%)",
    "MC Services RI Weight in All Items CPI (%)",
    "MC Services Weight in Core CPI (%)"
  ),
  t0 = c(
    round(recent_cpi$core_infl[4], 2),
    round(recent_cpi$ms_contrib[4], 2),
    round(recent_cpi$ms_contrib_share[4], 2),
    round(recent_cpi$ms_ri[4], 2),
    round(recent_cpi$ms_weight[4] * 100, 2)
  ),
  t1 = c(
    round(recent_cpi$core_infl[3], 2),
    round(recent_cpi$ms_contrib[3], 2),
    round(recent_cpi$ms_contrib_share[3], 2),
    round(recent_cpi$ms_ri[3], 2),
    round(recent_cpi$ms_weight[3] * 100, 2)
  ),
  t2 = c(
    round(recent_cpi$core_infl[2], 2),
    round(recent_cpi$ms_contrib[2], 2),
    round(recent_cpi$ms_contrib_share[2], 2),
    round(recent_cpi$ms_ri[2], 2),
    round(recent_cpi$ms_weight[2] * 100, 2)
  ),
  t3 = c(
    round(recent_cpi$core_infl[1], 2),
    round(recent_cpi$ms_contrib[1], 2),
    round(recent_cpi$ms_contrib_share[1], 2),
    round(recent_cpi$ms_ri[1], 2),
    round(recent_cpi$ms_weight[1] * 100, 2)
  ),
  avg = c(
    round(mean(trailing_12_cpi$core_infl), 2),
    round(mean(trailing_12_cpi$ms_contrib), 2),
    round(mean(trailing_12_cpi$ms_contrib_share), 2),
    round(mean(trailing_12_cpi$ms_ri), 2),
    round(mean(trailing_12_cpi$ms_weight) * 100, 2)
  )
)

names(cpi_contrib_tbl) <- c("Measure", col_labels_cpi)

knitr::kable(
  cpi_contrib_tbl,
  caption = "Medical Care Services Contribution to Core CPI Inflation (UIXFDG, UMS, RPCUMS, RUIXFDGM @CPIDATA)",
  align = c("l", rep("r", 5))
)
```

## Healthcare Sub-Component Price Indices

This section tracks price indices for key healthcare sub-components using PPI industry-based series and CPI dental services. Each row reports the most recent available price index level, month-over-month (MoM) change, and year-over-year (YoY) change, along with the component's expenditure weight (share of total PCE healthcare services spending). The two right-most columns both exclude the current month $t$ and use the trailing window ending at $t-1$: the average of the 12 monthly changes from $t-12$ to $t-1$, and the average of the 12 year-over-year changes from $t-12$ to $t-1$.

BEA uses these PPI and CPI prices as inputs to construct PCE healthcare sub-component price indices, so we pair each price series with its corresponding PCE nominal expenditure to compute weights. For dental services, BEA uses the CPI dental price in an intermediate step to derive the PCE dental price index, which is why we use PCE dental expenditures as the weight.

| Component | Price Index Code | Expenditure Weight Code | Description |
|------------------|------------------|------------------|------------------|
| Physicians | `R6211@PPIR` | `CSMPM@USNA` | PPI Offices of Physicians |
| Home Health | `R6216@PPIR` | `CSMOAM@USNA` | PPI Home Health Care Services |
| Medical & Diagnostic Labs | `R6215@PPIR` | `CSMOLM@USNA` | PPI Medical & Diagnostic Labs |
| Hospitals | `R622@PPIR` | `CSMHPM@USNA` | PPI Hospitals |
| Nursing Care | `R6231@PPIR` | `CSMHNM@USNA` | PPI Nursing Facilities |
| Dental Services | `UMSPT@CPIDATA` | `CSMDM@USNA` | CPI-U Dental Services |

For a given price index $P_t$:

$$\text{YoY}_t = \left(\frac{P_t}{P_{t-12}} - 1\right) \times 100, \qquad \text{MoM}_t = \left(\frac{P_t}{P_{t-1}} - 1\right) \times 100$$

The appended trailing columns are:

$$\text{Avg MoM, Last 12}_t = \frac{1}{12}\sum_{j=1}^{12}\left(\frac{P_{t-j}}{P_{t-j-1}} - 1\right)\times 100$$

$$\text{Avg YoY, Last 12}_t = \frac{1}{12}\sum_{j=1}^{12}\text{YoY}_{t-j}$$

Expenditure weights are each component's share of total healthcare services nominal expenditure:

$$w_{i,t} = \frac{NE_{i,t}}{NE_{hcserv,t}} \times 100$$

```{r hc-subcomponents}
# --------------------------------------------------------------------------
# Define sub-component series
# --------------------------------------------------------------------------
subcomp_specs <- tribble(
  ~label,                    ~pi_code,  ~pi_db,     ~ne_code,  ~ne_db,
  "Physicians",              "R6211",   "PPIR",     "csmpm",   "USNA",
  "Home Health",             "R6216",   "PPIR",     "csmoam",  "USNA",
  "Medical & Diagnostic Labs", "R6215", "PPIR",     "csmolm",  "USNA",
  "Hospitals",               "R622",    "PPIR",     "csmhpm",  "USNA",
  "Nursing Care",            "R6231",   "PPIR",     "csmhnm",  "USNA",
  "Dental Services",         "UMSPT",   "CPIDATA",  "csmdm",   "USNA"
)

# Pull total healthcare services NE for the denominator
hcserv_ne <- import_haver(series = "csdm@USNA", eop = FALSE) |>
  rename(hcserv_ne = csdm) |>
  arrange(date)

# --------------------------------------------------------------------------
# Build one row per sub-component
# --------------------------------------------------------------------------
build_subcomp_row <- function(label, pi_code, pi_db, ne_code, ne_db) {
  # Pull price index
  pi_full <- paste0(pi_code, "@", pi_db)
  pi_d <- import_haver(series = pi_full, eop = FALSE)
  pi_col <- setdiff(names(pi_d), "date")
  names(pi_d)[names(pi_d) == pi_col[1]] <- "pi"

  pi_d <- pi_d |>
    arrange(date) |>
    filter(!is.na(pi)) |>
    mutate(
      yoy = (pi / lag(pi, 12) - 1) * 100,
      mom = (pi / lag(pi, 1) - 1) * 100
    )

  # Pull expenditure weight
  ne_full <- paste0(ne_code, "@", ne_db)
  ne_d <- import_haver(series = ne_full, eop = FALSE)
  ne_col <- setdiff(names(ne_d), "date")
  names(ne_d)[names(ne_d) == ne_col[1]] <- "ne"

  ne_d <- ne_d |>
    arrange(date) |>
    filter(!is.na(ne))

  # Merge with total HC services NE to get weight
  ne_d <- ne_d |>
    inner_join(hcserv_ne, by = "date") |>
    mutate(weight = ne / hcserv_ne * 100)

  # Get latest complete observation (needs yoy, so at least 12 months of data)
  latest_pi <- pi_d |> filter(!is.na(yoy)) |> slice_tail(n = 1)
  latest_ne <- ne_d |> filter(date == latest_pi$date)
  recent_mom <- pi_d |>
    filter(!is.na(mom), date < latest_pi$date) |>
    slice_tail(n = 12)
  recent_yoy <- pi_d |>
    filter(!is.na(yoy), date < latest_pi$date) |>
    slice_tail(n = 12)

  # If weight date doesn't match, use most recent weight available

  if (nrow(latest_ne) == 0) {
    latest_ne <- ne_d |> slice_tail(n = 1)
  }

  data.frame(
    Component     = label,
    `As Of`       = format(latest_pi$date, "%b %Y"),
    `Weight (%)`  = round(latest_ne$weight[1], 1),
    `PI Level`    = round(latest_pi$pi, 2),
    `MoM (%)`     = round(latest_pi$mom, 2),
    `YoY (%)`     = round(latest_pi$yoy, 2),
    `Avg MoM, Last 12 (%)` = round(mean(recent_mom$mom, na.rm = TRUE), 2),
    `Avg YoY, Last 12 (%)` = round(mean(recent_yoy$yoy, na.rm = TRUE), 2),
    check.names   = FALSE
  )
}

subcomp_tbl <- purrr::pmap_dfr(subcomp_specs, build_subcomp_row)

knitr::kable(
  subcomp_tbl,
  caption = "Healthcare Sub-Component Price Indices with PCE Expenditure Weights",
  align = c("l", "l", rep("r", 6))
)
```

## Data Release

This section shows the most recent published release date for CPI, PPI, and PCE-related data, then lists all currently scheduled upcoming release dates available from the official agency calendars. CPI is published in the BLS **Consumer Price Index** release, PPI is published in the BLS **Producer Price Index** release, and PCE is published in BEA's **Personal Income and Outlays** release.

The latest published release date is pulled from FRED release history via `fredr::fredr_release_dates()`. The upcoming schedule is pulled from the official BLS and BEA release-calendar pages. If the FRED API key or the schedule pages are unavailable at render time, the section will display a short error message instead of stopping the full dashboard render.

```{r data-release, results='asis'}
release_specs <- tribble(
  ~series, ~release_id, ~agency, ~release_name, ~schedule_url, ~schedule_type,
  "CPI",   10L,         "BLS",   "Consumer Price Index",       "https://www.bls.gov/schedule/news_release/cpi.htm", "bls",
  "PPI",   46L,         "BLS",   "Producer Price Index",       "https://www.bls.gov/schedule/news_release/ppi.htm", "bls",
  "PCE",   54L,         "BEA",   "Personal Income and Outlays","https://www.bea.gov/news/schedule",                  "bea"
)

pull_latest_release <- function(series, release_id, agency, release_name, schedule_url, schedule_type) {
  last_error <- NULL

  for (attempt in 1:3) {
    out <- tryCatch(
      fredr::fredr_release_dates(
        release_id = release_id,
        sort_order = "desc",
        limit = 1
      ),
      error = function(e) {
        last_error <<- e
        NULL
      }
    )

    if (!is.null(out)) {
      return(
        out |>
          transmute(
            Series = series,
            Agency = agency,
            `News Release` = release_name,
            `Latest Release` = as.Date(date)
          )
      )
    }

    Sys.sleep(attempt)
  }

  stop(last_error)
}

pull_upcoming_bls <- function(series, agency, release_name, schedule_url) {
  tbl <- rvest::read_html(schedule_url) |>
    rvest::html_elements("table") |>
    purrr::pluck(2) |>
    rvest::html_table()

  tbl |>
    transmute(
      Series = series,
      Agency = agency,
      `News Release` = release_name,
      `Reference Period` = `Reference Month`,
      `Release Date` = as.Date(gsub("\\.", "", `Release Date`), format = "%b %d, %Y"),
      `Release Time` = `Release Time`
    )
}

pull_upcoming_bea <- function(series, agency, release_name, schedule_url) {
  tbl <- rvest::read_html(schedule_url) |>
    rvest::html_element("table") |>
    rvest::html_table()

  year_label <- names(tbl)[1]
  schedule_year <- sub(".*(20[0-9]{2}).*", "\\1", year_label)
  names(tbl) <- c("schedule_stamp", "release_type", "Release")[seq_along(names(tbl))]

  tbl |>
    transmute(
      schedule_stamp,
      release_entry = Release
    ) |>
    filter(grepl("^Personal Income and Outlays", release_entry)) |>
    mutate(
      Series = series,
      Agency = agency,
      `News Release` = release_name,
      `Reference Period` = trimws(sub("^Personal Income and Outlays,\\s*", "", release_entry)),
      `Release Date` = as.Date(
        paste(trimws(sub("\n.*$", "", schedule_stamp)), schedule_year),
        format = "%B %d %Y"
      ),
      `Release Time` = trimws(sub("^.*\n", "", schedule_stamp))
    ) |>
    select(Series, Agency, `News Release`, `Reference Period`, `Release Date`, `Release Time`)
}

latest_release <- tryCatch({
  purrr::pmap_dfr(release_specs, pull_latest_release)
}, error = function(e) e)

upcoming_release <- tryCatch({
  purrr::pmap_dfr(
    release_specs,
    function(series, release_id, agency, release_name, schedule_url, schedule_type) {
      if (schedule_type == "bls") {
        pull_upcoming_bls(series, agency, release_name, schedule_url)
      } else {
        pull_upcoming_bea(series, agency, release_name, schedule_url)
      }
    }
  )
}, error = function(e) e)

if (inherits(latest_release, "error") || inherits(upcoming_release, "error")) {
  err_msg <- if (inherits(latest_release, "error")) {
    conditionMessage(latest_release)
  } else {
    conditionMessage(upcoming_release)
  }

  cat(
    paste0(
      "Data release calendar unavailable: ",
      err_msg,
      ". Check that `FRED_API_KEY` is available and that the BLS/BEA schedule pages are reachable in the render environment."
    )
  )
} else {
  # Shift a "Month Year" (or bare "Month") reference period back one calendar
  # month. Used to infer the Latest Reference Period when the agency's
  # schedule page no longer lists the already-published release (BEA).
  shift_ref_period <- function(ref, fallback_year) {
    if (is.na(ref)) return(NA_character_)
    d <- suppressWarnings(lubridate::my(ref))
    if (is.na(d) && !is.na(fallback_year)) {
      d <- suppressWarnings(
        as.Date(paste("1", ref, fallback_year), format = "%d %B %Y")
      )
    }
    if (is.na(d)) return(NA_character_)
    format(d - months(1), "%B %Y")
  }

  full_schedule <- upcoming_release

  upcoming_release <- full_schedule |>
    inner_join(latest_release, by = c("Series", "Agency", "News Release")) |>
    filter(`Release Date` > `Latest Release`) |>
    arrange(Series, `Release Date`)

  latest_reference <- full_schedule |>
    inner_join(latest_release, by = c("Series", "Agency", "News Release")) |>
    filter(`Release Date` == `Latest Release`) |>
    select(
      Series, Agency, `News Release`,
      `Latest Reference Period` = `Reference Period`
    )

  release_summary <- latest_release |>
    left_join(latest_reference, by = c("Series", "Agency", "News Release")) |>
    left_join(
      upcoming_release |>
        group_by(Series, Agency, `News Release`) |>
        summarise(
          `Next Scheduled Release` = min(`Release Date`, na.rm = TRUE),
          `Next Reference Period` = `Reference Period`[which.min(`Release Date`)],
          `Upcoming Releases` = dplyr::n(),
          .groups = "drop"
        ),
      by = c("Series", "Agency", "News Release")
    ) |>
    mutate(
      `Latest Reference Period` = dplyr::coalesce(
        `Latest Reference Period`,
        purrr::map2_chr(
          `Next Reference Period`,
          lubridate::year(`Next Scheduled Release`),
          shift_ref_period
        )
      )
    ) |>
    select(
      Series,
      Agency,
      `News Release`,
      `Latest Release`,
      `Latest Reference Period`,
      `Next Scheduled Release`,
      `Next Reference Period`,
      `Upcoming Releases`
    ) |>
    mutate(
      `Latest Release` = format(`Latest Release`, "%b %d, %Y"),
      `Next Scheduled Release` = ifelse(
        is.na(`Next Scheduled Release`),
        NA_character_,
        format(`Next Scheduled Release`, "%b %d, %Y")
      )
    )

  upcoming_display <- upcoming_release |>
    transmute(
      Series,
      Agency,
      `News Release`,
      `Reference Period`,
      `Release Date` = format(`Release Date`, "%b %d, %Y"),
      `Release Time`
    )

  print(
    knitr::kable(
      release_summary,
      caption = "Latest Published Release and Next Scheduled Release",
      align = c("l", "l", "l", "r", "l", "r", "l", "r")
    )
  )

  cat("\n\n")

  print(
    knitr::kable(
      upcoming_display,
      caption = "Upcoming Release Calendar",
      align = c("l", "l", "l", "l", "r", "r")
    )
  )
}
```

## Data Staleness Monitor

This section checks, for every Haver series used across the project's PCE, PPI, and CPI code, how current the data is and whether any recent observations are missing.

Each series is pulled on its own. Within that pull, **t** is the most recent date Haver returns for the series. The "last 12 observations" used below are the 12 most recent months in that pull — the months **t−11 through t**. Every missing-data measure is anchored to this per-series window: it is judged against the series' own most recent Haver date, not against a shared calendar date or against the other series.

The table reports:

- **Source** — the data program the series belongs to (PCE, PPI, or CPI).
- **Code** — the Haver series mnemonic.
- **Database** — the Haver database the series is pulled from.
- **Description** — a plain-language description of the series.
- **Most Recent Haver Date** — the final date returned by the Haver pull (t), whether or not that date's value is missing.
- **Last Non-Missing Date** — the most recent date in the pull that has an actual (non-missing) value.
- **% Missing, Last 12 Obs** — the share of the 12 most recent observations (t−11 through t) whose value is missing.
- **Missing Dates, Last 12 Obs** — the specific months within that 12-observation window that are missing (blank if none are missing).
- **Used for PCE** — "Yes" if BEA uses the series as an input when constructing the PCE healthcare price indices, otherwise "No".

When the most recent Haver date is later than the last non-missing date, the series is returning empty (missing) values at the end of its pull. Rows are shaded by how far the last non-missing date trails the most recent month observed across all series in the table: green ("Current") matches that frontier, yellow ("Behind") is 1 month back, orange ("Delayed") is 2–5 months back, and red ("Stale") is 6 or more months back or missing entirely.

```{r staleness-monitor}
# --------------------------------------------------------------------------
# Define all Haver series used across the project, grouped by source
# --------------------------------------------------------------------------
series_registry <- tribble(
  ~series_code,   ~database,   ~source_group, ~description,

  # --- PCE Nominal Expenditures (USNA) --- (SA, Mil$). NE = nominal expenditures
  "cm",           "USNA",      "PCE",         "Overall PCE NE",
  "cxfem",        "USNA",      "PCE",         "Core PCE NE",
  "cgxfem",       "USNA",      "PCE",         "Goods ex F&E NE",
  "csm",          "USNA",      "PCE",         "Services NE",
  "csxhem",       "USNA",      "PCE",         "Services ex E&HC NE",
  "csdm",         "USNA",      "PCE",         "Healthcare Services NE",
  "csdom",        "USNA",      "PCE",         "Healthcare Services: Outpatient Services NE",
  "csmpm",        "USNA",      "PCE",         "Healthcare Services: Physician Services NE",
  "csmdm",        "USNA",      "PCE",         "Healthcare Services: Dental Services NE",
  "csmom",        "USNA",      "PCE",         "Healthcare Services: Paramedical Services NE",
  "csmnm",        "USNA",      "PCE",         "Healthcare Services: Hospital and Nursing Home Services NE",
  "csmolm",       "USNA",      "PCE",         "Healthcare Services: Paramedical: Medical Laboratories NE",
  "csmoom",       "USNA",      "PCE",         "Healthcare Services: Paramedical: Other Professional Medical Services NE",
  "csmhpm",       "USNA",      "PCE",         "Healthcare Services: Hospital Services NE",
  "csmhnm",       "USNA",      "PCE",         "Healthcare Services: Nursing Homes NE",
  "ctgm",         "USNA",      "PCE",         "Total Goods NE",
  # "csmoa",        "USNA",      "PCE",         "Healthcare Services: Paramedical Services", # last published value was as of 2025-01-01
  "csmoam",       "USNA",      "PCE",         "Healthcare Services: Paramedical: Home Health Care NE (monthly)",

  # --- PCE Price Indices (USNA) --- (seasonally adjusted= SA)
  "jcm",          "USNA",      "PCE",         "Overall PCE PI",
  "jcxfem",       "USNA",      "PCE",         "Core PCE PI (ex F&E)",
  "jcgxfem",      "USNA",      "PCE",         "Goods ex F&E PCE PI",
  "jcsm",         "USNA",      "PCE",         "Services PCE PI",
  "jcsxhem",      "USNA",      "PCE",         "Services ex E&HC PCE PI",
  "jcsdm",        "USNA",      "PCE",         "Healthcare Services PCE PI",
  "jcsdom",       "USNA",      "PCE",         "Outpatient Services PCE PI",
  "jcsmpm",       "USNA",      "PCE",         "Physician Services PCE PI",
  "jcsmdm",       "USNA",      "PCE",         "Dental Services PCE PI",
  "jcsmom",       "USNA",      "PCE",         "Paramedical Services PCE PI",
  "jcsmnm",       "USNA",      "PCE",         "Hospital & Nursing PCE PI",
  "jcsmhpm",      "USNA",      "PCE",         "Hospital Services PCE PI",
  "jcsmhnm",      "USNA",      "PCE",         "Nursing Homes PCE PI",
  "jctgm",        "USNA",      "PCE",         "Total Goods PCE PI",

  # --- Core PCE (USECON) --- (SA)
  "jcxfebm",      "USECON",    "PCE",         "Core PCE PI (USECON)",

  # --- PPI Industry-Based (PPIR) --- (appear to be SA)
  "RHC",          "PPIR",      "PPI",         "PPI Selected Health Care Industries",
  "R6211",        "PPIR",      "PPI",         "PPI Offices of Physicians",
  "R6212",        "PPIR",      "PPI",         "PPI Offices of Dentists",
  "R6215",        "PPIR",      "PPI",         "PPI Medical & Diagnostic Labs",
  "R6216",        "PPIR",      "PPI",         "PPI Home Health",
  "R621991",      "PPIR",      "PPI",         "PPI Blood and Organ Banks",
  "R622",         "PPIR",      "PPI",         "PPI Hospitals",
  "R6231",        "PPIR",      "PPI",         "PPI Nursing Facilities",
  "R62321",       "PPIR",      "PPI",         "PPI Residential Developmental Disability Homes",

  # special index (table 11) (NSA)
  "R62211A2",     "PPIR",      "PPI",         "PPI Hospitals - Medicare",
  "R62211A4",     "PPIR",      "PPI",         "PPI Hospitals - Medicaid",
  "R62211A6",     "PPIR",      "PPI",         "PPI Hospitals - Private/Other",

  # --- PPI Commodity-Based (PPI) --- (only pulling NSA because it's the most available)
  "P51",          "PPI",       "PPI",         "PPI Healthcare Services (Commodity)",
  "P511",         "PPI",       "PPI",         "PPI Healthcare Services: Outpatient (Commodity)",
  "P511104B",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Medicare (Commodity)",
  "P511104C",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Medicaid (Commodity)",
  "P511104D",     "PPI",       "PPI",         "PPI Healthcare Services: Outpatient: Hospital: Private/All Other (Commodity)",

  "P512",         "PPI",       "PPI",         "PPI Healthcare Services: Inpatient (Commodity)",
  "P512101B",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Medicare (Commodity)",
  "P512101C",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Medicaid (Commodity)",
  "P512101D",     "PPI",       "PPI",         "PPI Healthcare Services: Inpatient: Hospital: Private/All Other (Commodity)",

  # --- CPI (CPIDATA) ---
  "UM",           "CPIDATA",   "CPI",         "CPI-U Medical Care (SA)",
  "UMC",          "CPIDATA",   "CPI",         "CPI-U Medical Care Commodities (SA)",
  "UMG",          "CPIDATA",   "CPI",         "CPI-U Medicinal Drugs (SA)",
  "UMP",          "CPIDATA",   "CPI",         "CPI-U Prescription Drugs (SA)",
  "UMGNA",        "CPIDATA",   "CPI",         "CPI-U Nonprescription Drugs (SA)",
  "UMS",          "CPIDATA",   "CPI",         "CPI-U Medicare Services (SA)",
  "UMSPP",        "CPIDATA",   "CPI",         "CPI-U Physicians' Services (SA)",
  "UMSPT",        "CPIDATA",   "CPI",         "CPI-U Dental Services (SA)",
  "UMSPE",        "CPIDATA",   "CPI",         "CPI-U Eyeglasses & Eyecare (SA)",
  "UMSPS",        "CPIDATA",   "CPI",         "CPI-U Services by Other Medical Professionals (SA)",
  "UMSH",         "CPIDATA",   "CPI",         "CPI-U Hospital and Related Services (SA)",
  "UMSHS",        "CPIDATA",   "CPI",         "CPI-U Hospital Services (SA)",
  "UMSHSI",       "CPIDATA",   "CPI",         "CPI-U Inpatient Hospital Services (SA)",
  "UMSHSO",       "CPIDATA",   "CPI",         "CPI-U Outpatient Hospital Services (SA)",
  "UMSNS",        "CPIDATA",   "CPI",         "CPI-U Nursing Homes and Adult Daycare (SA)",
  # Not seasonally adjusted
  "UMN",          "CPIDATA",   "CPI",         "CPI-U Medical Care (NSA)",
  "UMCN",         "CPIDATA",   "CPI",         "CPI-U Medical Care Commodities (NSA)",
  "UMGN",         "CPIDATA",   "CPI",         "CPI-U Medicinal Drugs (NSA)",
  "UMPN",         "CPIDATA",   "CPI",         "CPI-U Prescription Drugs (NSA)",
  "UMGNN",        "CPIDATA",   "CPI",         "CPI-U Nonprescription Drugs (NSA)",
  "UMQN",         "CPIDATA",   "CPI",         "CPI-U Medical Equipment and Supplies (NSA)",
  "UMSN",         "CPIDATA",   "CPI",         "CPI-U Medical Care Services (NSA)",
  "UMSPN",        "CPIDATA",   "CPI",         "CPI-U Professional Medical Care Services (NSA)",
  "UMSPPN",       "CPIDATA",   "CPI",         "CPI-U Physicians' Services (NSA)",
  "UMSPTN",       "CPIDATA",   "CPI",         "CPI-U Dental Services (NSA)",
  "UMSPEN",       "CPIDATA",   "CPI",         "CPI-U Eyeglasses and Eye Care (NSA)",
  "UMSPSN",       "CPIDATA",   "CPI",         "CPI-U Services by Other Medical Professionals (NSA)",
  "UMSHN",        "CPIDATA",   "CPI",         "CPI-U Hospital and Related Services (NSA)",
  "UMSHSN",       "CPIDATA",   "CPI",         "CPI-U Hospital Services (NSA)",
  "UMSHSIN",      "CPIDATA",   "CPI",         "CPI-U Inpatient Hospital Services (NSA)",
  "UMSHSON",      "CPIDATA",   "CPI",         "CPI-U Outpatient Hospital Services (NSA)",
  "UMSNSN",       "CPIDATA",   "CPI",         "CPI-U Nursing Homes and Adult Daycare (NSA)",
  "UOEECN",       "CPIDATA",   "CPI",         "CPI-U Home Health Care",
  "UMSIN",        "CPIDATA",   "CPI",         "CPI-U Health Insurance",
  "UIXFDG",       "CPIDATA",   "CPI",         "CPI-U All Items Less Food & Energy (SA)",
  "RPCUMS",       "CPIDATA",   "CPI",         "CPI-U RI Weight: Medical Care Services",
  "RUIXFDGM",     "CPIDATA",   "CPI",         "CPI-U RI Weight: All Items Less Food & Energy"
)

# --------------------------------------------------------------------------
# Pull each series, extract the last Haver date and last non-NA date, and
# compute the share of missing values in the most recent 12 observations.
# --------------------------------------------------------------------------
get_series_status <- function(code, db) {
  full_code <- paste0(code, "@", db)
  tryCatch({
    d <- import_haver(series = full_code, eop = FALSE)
    col <- setdiff(names(d), "date")
    d <- d |>
      arrange(date)

    last_haver_date <- if (nrow(d) == 0) {
      as.Date(NA)
    } else {
      max(d$date)
    }

    d_nonmissing <- d |> filter(!is.na(.data[[col[1]]]))
    last_observed_date <- if (nrow(d_nonmissing) == 0) {
      as.Date(NA)
    } else {
      max(d_nonmissing$date)
    }

    recent_12 <- d |> slice_tail(n = 12)
    pct_na_last_12 <- if (nrow(recent_12) == 0) {
      NA_real_
    } else {
      mean(is.na(recent_12[[col[1]]])) * 100
    }

    # Dates among the most recent 12 returned observations whose value is missing
    missing_dates_last_12 <- if (nrow(recent_12) == 0) {
      NA_character_
    } else {
      miss <- recent_12$date[is.na(recent_12[[col[1]]])]
      if (length(miss) == 0) {
        ""
      } else {
        paste(format(miss, "%b %Y"), collapse = ", ")
      }
    }

    tibble(
      last_haver_date = last_haver_date,
      last_observed_date = last_observed_date,
      pct_na_last_12 = pct_na_last_12,
      missing_dates_last_12 = missing_dates_last_12
    )
  }, error = function(e) {
    tibble(
      last_haver_date = as.Date(NA),
      last_observed_date = as.Date(NA),
      pct_na_last_12 = NA_real_,
      missing_dates_last_12 = NA_character_
    )
  })
}

# Series BEA uses as inputs when constructing the PCE healthcare price indices
pce_release_series <- c(
  "R6211@PPIR", "R6216@PPIR", "R6215@PPIR", "R622@PPIR", "R6231@PPIR",
  "UMSPT@CPIDATA", "UMSPTN@CPIDATA", "UMSPS@CPIDATA", "UMSPSN@CPIDATA"
)

staleness <- series_registry |>
  mutate(
    series_status = purrr::map2(series_code, database, get_series_status),
    in_pce_release = ifelse(
      paste0(series_code, "@", database) %in% pce_release_series,
      "Yes", "No"
    )
  ) |>
  tidyr::unnest(series_status)

# --------------------------------------------------------------------------
# Determine the most recent end date and flag stale series
# --------------------------------------------------------------------------
most_recent <- max(staleness$last_observed_date, na.rm = TRUE)

staleness <- staleness |>
  mutate(
    months_behind = as.numeric(
      (year(most_recent) - year(last_observed_date)) * 12 +
      (month(most_recent) - month(last_observed_date))
    ),
    status = case_when(
      is.na(last_observed_date) ~ "Stale",
      months_behind == 0    ~ "Current",
      months_behind == 1    ~ "Behind",
      months_behind <= 5    ~ "Delayed",
      TRUE                  ~ "Stale"
    )
  )

# --------------------------------------------------------------------------
# Single color-coded table for all series
# --------------------------------------------------------------------------
staleness_display <- staleness |>
  mutate(
    `NA Share, Last 12 Obs (%)` = round(pct_na_last_12, 1),
    `Missing Dates, Last 12 Obs` = missing_dates_last_12
  ) |>
  select(
    source_group,
    series_code,
    database,
    description,
    last_haver_date,
    last_observed_date,
    `NA Share, Last 12 Obs (%)`,
    `Missing Dates, Last 12 Obs`,
    in_pce_release,
    status
  ) |>
  arrange(status, source_group, series_code)

DT::datatable(
  staleness_display,
  caption = paste0(
    "Series Staleness Monitor (most recent observed: ",
    format(most_recent, "%B %Y"), ")"
  ),
  colnames = c(
    "Source",
    "Code",
    "Database",
    "Description",
    "Most Recent Haver Date",
    "Last Non-Missing Date",
    "% Missing, Last 12 Obs",
    "Missing Dates, Last 12 Obs",
    "Used for PCE",
    "Status"
  ),
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    scrollX = TRUE,
    scrollY = TRUE,
    dom = 'Blfrtip',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 10,
    lengthMenu = list(c(10, 25, 50, -1), c('10', '25', '50', 'All'))
  )
) |>
  DT::formatStyle(
    "status",
    target = "row",
    backgroundColor = DT::styleEqual(
      c("Current", "Behind", "Delayed", "Stale"),
      c("#d4edda", "#fff3cd", "#ffd6a5", "#f8d7da")
    )
  )
```
