---
title: 'Inflation Dashboard'
author: 'Shria Holla, Kelli Marquardt, and Aryan Safi'
date: today
format: html
editor: visual
code-fold: true
execute: 
  warning: false
  message: false
toc: true
embed-resources: true
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(
  fig.width = 12,   # width in inches
  fig.height = 6,  # height in inches
  echo = TRUE,     # show code by default
  warning = FALSE, # hide warnings
  message = FALSE  # hide messages
)
# read in the population data from ipums
pacman::p_load(here,
               tidyverse,
               data.table,
               importhaver,
               Haver,
               readxl,
               fredr)

# set theme 
theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(size = 22, face = 'bold'),
      plot.subtitle = element_text(size = 20),
      legend.title = element_text(size = 16),
      legend.text = element_text(size = 14),
      legend.position = 'bottom',
      axis.text = element_text(size = 13),
      axis.title = element_text(size = 18),
      plot.caption = element_text(size = 12),
      axis.line = element_line(color = "black"),
      strip.text = element_text(size = 16),
      axis.ticks = element_line(size = 1.01)
    )
)

```


## Data Staleness Monitor

This section checks the last available date for every Haver series used across the project's PCE, PPI, and CPI code. Rather than listing each series individually, it summarizes by distinct end-date levels and flags any series that is 3 or more months behind the most recent data date.

```{r staleness-monitor}
# --------------------------------------------------------------------------
# Define all Haver series used across the project, grouped by source
# --------------------------------------------------------------------------
series_registry <- tribble(
  ~series_code,   ~database,   ~source_group, ~description,
  
  # --- PCE Nominal Expenditures (USNA) ---
  "cm",           "USNA",      "PCE",         "Overall PCE NE",
  "cxfem",        "USNA",      "PCE",         "Core PCE NE",
  "cgxfem",       "USNA",      "PCE",         "Goods ex F&E NE",
  "csm",          "USNA",      "PCE",         "Services NE",
  "csxhem",       "USNA",      "PCE",         "Services ex E&HC NE",
  "csdm",         "USNA",      "PCE",         "HC Services NE",
  "csdom",        "USNA",      "PCE",         "Outpatient Services NE",
  "csmpm",        "USNA",      "PCE",         "Physician Services NE",
  "csmdm",        "USNA",      "PCE",         "Dental Services NE",
  "csmom",        "USNA",      "PCE",         "Paramedical Services NE",
  "csmnm",        "USNA",      "PCE",         "Home & Nursing NE",
  "csmhpm",       "USNA",      "PCE",         "Hospital Services NE",
  "csmhnm",       "USNA",      "PCE",         "Nursing Homes NE",
  "ctgm",         "USNA",      "PCE",         "Total Goods NE",
  
  # --- PCE Price Indices (USNA) ---
  "jcm",          "USNA",      "PCE",         "Overall PCE PI",
  "jcxfem",       "USNA",      "PCE",         "Core PCE PI",
  "jcgxfem",      "USNA",      "PCE",         "Goods ex F&E PI",
  "jcsm",         "USNA",      "PCE",         "Services PI",
  "jcsxhem",      "USNA",      "PCE",         "Services ex E&HC PI",
  "jcsdm",        "USNA",      "PCE",         "HC Services PI",
  "jcsdom",       "USNA",      "PCE",         "Outpatient Services PI",
  "jcsmpm",       "USNA",      "PCE",         "Physician Services PI",
  "jcsmdm",       "USNA",      "PCE",         "Dental Services PI",
  "jcsmom",       "USNA",      "PCE",         "Paramedical Services PI",
  "jcsmnm",       "USNA",      "PCE",         "Home & Nursing PI",
  "jcsmhpm",      "USNA",      "PCE",         "Hospital Services PI",
  "jcsmhnm",      "USNA",      "PCE",         "Nursing Homes PI",
  "jctgm",        "USNA",      "PCE",         "Total Goods PI",
  
  # --- Core PCE (USECON) ---
  "jcxfebm",      "USECON",    "PCE",         "Core PCE PI (USECON)",
  
  # --- PPI Industry-Based (PPIR) ---
  "r6211",        "PPIR",      "PPI",         "PPI Offices of Physicians",
  "r6215",        "PPIR",      "PPI",         "PPI Medical Labs",
  "r6216",        "PPIR",      "PPI",         "PPI Home Health",
  "r622",         "PPIR",      "PPI",         "PPI Hospitals",
  "r6231",        "PPIR",      "PPI",         "PPI Nursing Facilities",
  "R62211A2",     "PPIR",      "PPI",         "PPI Hospitals - Medicare",
  "R62211A4",     "PPIR",      "PPI",         "PPI Hospitals - Medicaid",
  "R62211A6",     "PPIR",      "PPI",         "PPI Hospitals - Private/Other",
  
  # --- PPI Commodity-Based (PPI) ---
  "P512101B",     "PPI",       "PPI",         "PPI Hospital Inpatient (Commodity)",
  "P512101C",     "PPI",       "PPI",         "PPI Hospital Inpatient Medicare (Commodity)",
  "P512101D",     "PPI",       "PPI",         "PPI Hospital Inpatient Medicaid (Commodity)",
  "P511104B",     "PPI",       "PPI",         "PPI Hospital Outpatient (Commodity)",
  "P511104C",     "PPI",       "PPI",         "PPI Hospital Outpatient Medicare (Commodity)",
  "P511104D",     "PPI",       "PPI",         "PPI Hospital Outpatient Medicaid (Commodity)",
  
  # --- CPI (CPIDATA) ---
  "umsptn",       "CPIDATA",   "CPI",         "CPI Medical Care Services (NSA)",
  "umspsn",       "CPIDATA",   "CPI",         "CPI Hospital Services (NSA)"
)

DT::datatable(
  series_registry,
  caption = "Haver Codes",
  colnames = c("Series Code", "Database", "Source Group", "Description"),
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    # Added 'l' to dom to show the length menu
    dom = 'Blfrtip', 
    buttons = c('copy', 'csv', 'excel'),
    # Sets the default number of rows to display
    pageLength = 10, 
    # Defines the options available in the dropdown
    lengthMenu = list(c(10, 25, 50, -1), c('10', '25', '50', 'All'))
  )
)

```


```{r}
# --------------------------------------------------------------------------
# Pull each series and extract the last non-NA date
# --------------------------------------------------------------------------
get_end_date <- function(code, db) {
  full_code <- paste0(code, "@", db)
  tryCatch({
    d <- import_haver(series = full_code, eop = FALSE) |> 
      na.omit()
    col <- setdiff(names(d), "date")
    d <- d |> filter(!is.na(.data[[col[1]]]))
    if (nrow(d) == 0) return(as.Date(NA))
    max(d$date)
  }, error = function(e) as.Date(NA))
}

staleness <- series_registry |>
  rowwise() |>
  mutate(end_date = get_end_date(series_code, database)) |>
  ungroup()

# --------------------------------------------------------------------------
# Determine the most recent end date and flag stale series
# --------------------------------------------------------------------------
most_recent <- max(staleness$end_date, na.rm = TRUE)
stale_threshold <- most_recent %m-% months(3)

staleness <- staleness |>
  mutate(
    months_behind = as.numeric(
      difftime(most_recent, end_date, units = "days")
    ) / 30.44,
    flag = case_when(
      is.na(end_date)            ~ "MISSING",
      end_date <= stale_threshold ~ "STALE",
      TRUE                       ~ "OK"
    )
  )
```

### End-Date Summary

```{r staleness-summary}
# Summary table: how many series end at each date, by source group
end_date_summary <- staleness |>
  group_by(source_group, end_date) |>
  summarise(
    n_series = n(),
    series = paste(series_code, collapse = ", "),
    .groups = "drop"
  ) |>
  arrange(source_group, desc(end_date))

DT::datatable(
  end_date_summary,
  caption = "Series Count by End Date and Source",
  colnames = c("Source", "End Date", "# Series", "Series Codes"),
  rownames = FALSE,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel'),
    pageLength = 25
  )
)
```

### Stale or Missing Series

Series that are 3+ months behind the most recent data date (`r format(most_recent, "%B %Y")`) or that failed to load.

```{r staleness-flagged}
flagged <- staleness |>
  filter(flag != "OK") |>
  select(source_group, series_code, database, description, end_date, months_behind, flag) |>
  arrange(flag, desc(months_behind))

if (nrow(flagged) == 0) {
  cat("All series are current -- no investigation needed.")
} else {
  DT::datatable(
    flagged,
    caption = paste0("Series requiring investigation (threshold: ",
                     format(stale_threshold, "%B %Y"), ")"),
    colnames = c("Source", "Code", "Database", "Description",
                 "End Date", "Months Behind", "Status"),
    rownames = FALSE,
    extensions = 'Buttons',
    options = list(
      dom = 'Bfrtip',
      buttons = c('copy', 'csv', 'excel')
    )
  ) |>
    DT::formatRound("months_behind", digits = 1) |>
    DT::formatStyle(
      "flag",
      backgroundColor = DT::styleEqual(
        c("STALE", "MISSING"),
        c("#fff3cd", "#f8d7da")
      )
    )
}
```

