---
title: "Healthcare Graphs"
author: "Shria Holla"
output: html_document
date: "`Sys.Date()"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())
packages <- c("devtools", "importhaver", "ggplot2", "dplyr", "tidyr", "Haver", "foreign")
lapply(packages, library, character.only = TRUE)
```

```{r}
source("hcgraphs_functions.R")
```



##################################################################################################################################################################################################################################

## (1) Overall PCE, Core PCE, and Healthcare Services Inflation Over Time
```{r}

#Pull Data
usnalist <- c("jcm", "jcxfem", "jcsdm")
rename_list <- c("ov_pi", "core_pi", "hcserv_pi")
g1data <- get_haver_data(usnalist, rename_list = rename_list)

#Calculate Inflation
components <- c("ov", "core", "hcserv")
g1data <- calc_infl(g1data)

```
### Plot Graphs
```{r}
g1plotspecs <- data.frame(
  title = c(
    "Overall PCE, Core PCE, and Healthcare Services Inflation",
    "Overall PCE, Core PCE, and Healthcare Services Inflation since 2000"
  ),
  filename = c(
    "ov_core_hcserv_infl.png",
    "ov_core_hcserv_infl_2000.png"
  ),
  start_date = c(
    NA, 
    "2000-01-31"),
  stringsAsFactors = FALSE
)

for (i in 1:nrow(g1plotspecs)) {
  
  g1 <- plot_line(
    data = g1data,
    y_cols = c(
      "Overall PCE" = "ov_infl",
      "Core PCE" = "core_infl", 
      "Healthcare Services" = "hcserv_infl"
      ),
    colors = c(
      "Overall PCE" = "darkgreen", 
      "Core PCE" = "deeppink", 
      "Healthcare Services" = "blue"
      ),
    title = g1plotspecs$title[i],
    y_label = "Percent (%)",
    show_legend = FALSE,
    start_date = if(!is.na(g1plotspecs$start_date[i])) as.Date(g1plotspecs$start_date[i]) else NULL
    )
  
  print(g1)
  save_plot(g1, g1plotspecs$filename[i])
}

```
 
### Clean Up G1 Values
```{r}
rm(g1, g1data, g1plotspecs, components)
```










##################################################################################################################################################################################################################################

## (2) Weight of Healthcare Services in Overall PCE
```{r}

#Pull Data
usnalist <- c("cm", "csdm")
rename_list <- c("ov_ne", "hcserv_ne")
g2data <- get_haver_data(usnalist, rename_list = rename_list)

#Calculate Expenditure Share
components <- c("hcserv")
g2data <- g2data %>%
  calc_shares(aggvar = "ov") %>%
  mutate(hcserv_expshare = hcserv_expshare*100)

```
### Graph Weight of Healthcare Services in Overall PCE
```{r}
# Create healthcare weight graph using plot_line function with automatic features
g2 <- plot_line(
  data = g2data,
  y_cols = list("hcserv weight in ov" = "hcserv_expshare"),
  colors = c("hcserv weight in ov" = "blue"),
  title = "Weight of Healthcare Services in Overall PCE",
  y_label = "Percent (%) Share",
  show_legend = FALSE
)

print(g2)
save_plot(g2, "hcserv_weight_ov.png")
```


### Graph YoY Change in Healthcare Services Weight in Overall PCE, Monthly
```{r}
#Find year over year change in weight
g2myoy <- calc_infl(g2data, pi_suffix = "_expshare", infl_suffix = "_weight_yoy")

# Assign Plot Specifications
g2mplotspecs <- data.frame(
  title = c(
    "YoY Change in Weight of Healthcare Services in Overall PCE\nMonthly",
    "YoY Change in Weight of Healthcare Services in Overall PCE since 2000\nMonthly"
  ),
  filename = c(
    "hcserv_weight_yoy_ov.png",
    "hcserv_weight_yoy_ov_2000.png"
  ),
  start_date = c(
    NA, 
    "2000-01-31"
    ),
  stringsAsFactors = FALSE
)

#Graph
for (i in 1:nrow(g2mplotspecs)) {
  
  g2m <- plot_line(
    data = g2myoy,
    y_cols = c("weight change" = "hcserv_weight_yoy"),
    colors = c("weight change" = "blue"),
    title = g2mplotspecs$title[i],
    y_label = "Percent (%) Share",
    show_legend = FALSE,
    start_date = if(!is.na(g2mplotspecs$start_date[i])) as.Date(g2mplotspecs$start_date[i]) else NULL
  )
  
  print(g2m)
  save_plot(g2m, g2mplotspecs$filename[[i]])
}

```

### Graph YoY Change in Healthcare Services Weight in Overall PCE, Annual
```{r}
#Annual
g2ayoy <- g2data %>%
  calc_infl(pi_suffix = "_expshare", infl_suffix = "_weight_yoy") %>%
  mutate(year = format(date, "%Y")) %>%
  group_by(year) %>%
  summarize(yr_hcserv_weight_yoy = mean(hcserv_weight_yoy))

# Assign Plot Specifications
g2aplotspecs <- data.frame(
  title = c(
    "YoY Change in Weight of Healthcare Services in Overall PCE\nAnnual",
    "YoY Change in Weight of Healthcare Services in Overall PCE since 2000\nAnnual"
  ),
  filename = c(
    "hcserv_weight_yoy_ov_annual.png",
    "hcserv_weight_yoy_ov_2000_annual.png"
  ),
  start_date = c(
    NA, 
    2000
    ),
  stringsAsFactors = FALSE
)

# Graph
for (i in 1:nrow(g2aplotspecs)) {
  g2a <- plot_line_year(
    data = g2ayoy,
    x_col = "year",
    y_cols = list("YrlyYOY" = "yr_hcserv_weight_yoy"),
    colors = c("YrlyYOY" = "blue"),
    title = g2aplotspecs$title[i],
    y_label = "Percent (%)",
    show_legend = FALSE,
    start_date = if (is.na(g2aplotspecs$start_date[i])) NULL else g2aplotspecs$start_date[i]
    )
  print(g2a)
  save_plot(g2a, g2aplotspecs$filename[[i]])
}
```




### Clean Up G2 Values
```{r}
rm(g2, g2data, g2aplotspecs, components, g2a, g2m, g2mplotspecs, g2ayoy, g2myoy)
```



##################################################################################################################################################################################################################################

## (3) Weight of Healthcare Subcomponents in Healthcare Services

```{r}
# Pull Data
usnalist <- c("csdm", "csdom", "csmpm", "csmdm", "csmom", "csmnm", "csmhpm", "csmhnm")
rename_list <- c("hcserv_ne", "opserv_ne", "physerv_ne", "dentserv_ne", "paraserv_ne", "hnserv_ne", "hosp_ne", "nurs_ne")
g3data <- get_haver_data(usnalist, rename_list = rename_list)
```





##################################################################################################################################################################################################################################


### Graph Total Expenditure on Healthcare Services Subcomponents in Dollars
```{r}
# Levels Data
g3levels <- g3data %>%
 mutate(across(ends_with("ne"), ~.x/1000000))

# Graph
g3plotspecs <- data.frame(
  title = c(
    "Household Expenditure on Healthcare Services Subcomponents",
    "Household Expenditure on Healthcare Services Subcomponents since 2000"
  ),
  filename = c(
    "subcomp_exp_hcserv.png",
    "subcomp_exp_hcserv_2000.png"
  ),
  start_date = c(
    NA, 
    "2000-01-31"
    ),
  stringsAsFactors = FALSE
)

for (i in 1:nrow(g3plotspecs)) {
  
  g3lev <- plot_line(
    data = g3levels,
    y_cols = list(
      "Physician" = "physerv_ne",
      "Dental" = "dentserv_ne",
      "Paramedical" = "paraserv_ne", 
      "Hospital" = "hosp_ne",
      "Nursing Homes" = "nurs_ne"
      ),
    colors = c(
      "Physician" = "darkblue",
      "Dental" = "forestgreen", 
      "Paramedical" = "darkorange",
      "Hospital" = "purple",
      "Nursing Homes" = "firebrick"
    ),
    title = g3plotspecs$title[i],
    y_label = "Millions of Dollars",
    start_date = if(!is.na(g3plotspecs$start_date[i])) as.Date(g3plotspecs$start_date[i]) else NULL
  )
  
  print(g3lev)
  save_plot(g3lev, g3plotspecs$filename[i])
}

```

### Graph Weight of Subcomponents in Healthcare Services
```{r}

#Calculate Shares
components <- c("physerv", "dentserv", "paraserv", "hosp", "nurs")
g3shares <- g3data %>%
  calc_shares(aggvar = "hcserv") %>%
  mutate(across(ends_with("expshare"), ~.x*100))

# Graph
g3_share <- plot_line(
  data = g3shares,
  y_cols = list(
    "Physician" = "physerv_expshare",
    "Dental" = "dentserv_expshare",
    "Paramedical" = "paraserv_expshare", 
    "Hospital" = "hosp_expshare",
    "Nursing Homes" = "nurs_expshare"
  ),
  colors = c(
    "Physician" = "darkblue",
    "Dental" = "forestgreen", 
    "Paramedical" = "darkorange",
    "Hospital" = "purple",
    "Nursing Homes" = "firebrick"
  ),
  title = "Weight of Subcomponents in Healthcare Services",
  y_label = "Percent (%) Share"
)

# Display the plot
print(g3_share)
save_plot(g3_share, "subcomp_weight_hcserv.png")

```

### Graph YoY Change in Subcomponent Weights in Healthcare Services
```{r}

#Calculate YoY Change in Expenditure Shares
g3wtchng <- calc_infl(g3shares, pi_suffix = "_expshare", infl_suffix = "_yoy")

#Graph
g3_yoywt <- plot_line(
  data = g3wtchng,
  y_cols = list(
    "Physician" = "physerv_yoy",
    "Dental" = "dentserv_yoy",
    "Paramedical" = "paraserv_yoy", 
    "Hospital" = "hosp_yoy",
    "Nursing Homes" = "nurs_yoy"
  ),
  colors = c(
    "Physician" = "darkblue",
    "Dental" = "forestgreen", 
    "Paramedical" = "darkorange",
    "Hospital" = "purple",
    "Nursing Homes" = "firebrick"
  ),
  title = "YoY Change in Subcomponent Shares of Healthcare Services Expenditures",
  y_label = "Percent (%) Share",
  start_date = as.Date("2000-01-31")
)

print(g3_yoywt)
save_plot(g3_yoywt, "subcomp_weight_yoy_hcserv_2000.png")
```

### Graph YoY Change in Weights of Subcomponents in Healthcare Services, Limited y-axis
```{r}
#Same graph with limited yaxis limits
g3_yoywt_limit <- plot_line(
  data = g3wtchng,
  y_cols = list(
    "Physician" = "physerv_yoy",
    "Dental" = "dentserv_yoy",
    "Paramedical" = "paraserv_yoy", 
    "Hospital" = "hosp_yoy",
    "Nursing Homes" = "nurs_yoy"
  ),
  colors = c(
    "Physician" = "darkblue",
    "Dental" = "forestgreen", 
    "Paramedical" = "darkorange",
    "Hospital" = "purple",
    "Nursing Homes" = "firebrick"
  ),
  title = "YoY Change in Subcomponent Shares of Healthcare Services Expenditures",
  y_label = "Percent (%) Share",
 start_date = as.Date("2000-01-31"),
 y_limits = c(-10, 10),
 y_breaks = c(-10, -8, -6, -4, -2, 0, 2, 4, 6, 8, 10)
)

print(g3_yoywt_limit)
save_plot(g3_yoywt_limit, "subcomp_weight_yoy_hcserv_2000_limit.png")


```

### Clean Up G3 Values
```{r}
rm(g3data, g3levels, g3plotspecs, g3shares, g3wtchng, g3_share, g3_yoywt, g3_yoywt_limit, g3lev, components)
```

##################################################################################################################################################################################################################################
## (4) Healthcare Services Inflation Decomposition

```{r}

# Pull In Data
usnalist <- c("csdm", "csdom", "csmpm", "csmdm", "csmom", "csmnm", "csmhpm", "csmhnm", "jcsdm", "jcsdom", "jcsmpm", "jcsmdm", "jcsmom", "jcsmnm", "jcsmhpm", "jcsmhnm")
rename_list <- c("hcserv_ne", "opserv_ne", "physerv_ne", "dentserv_ne", "paraserv_ne", "hnserv_ne", "hosp_ne", "nurs_ne", "hcserv_pi", "opserv_pi", "physerv_pi", "dentserv_pi", "paraserv_pi", "hnserv_pi", "hosp_pi", "nurs_pi")
g4data <- get_haver_data(usnalist, rename_list = rename_list)

#Find inflation, shares, and contributions
components <- c("physerv", "dentserv", "paraserv", "hosp", "nurs")
g4data <- g4data %>%
  calc_infl() %>%
  calc_shares(aggvar = "hcserv") %>%
  calc_contrib() %>%
  calc_aggregate(aggvar = "hcserv")
```

##################################################################################################################################################################################################################################

```{r}
# components <- "physerv"
# testdataset <- g4data %>%
#   mutate(
#     physerv_pi_lag12 = lag(physerv_pi, 12),
#     physerv_ne_lag12 = lag(physerv_ne, 12),
#     pricegrowth = (physerv_pi - physerv_pi_lag12) / physerv_pi_lag12,
#     expgrowth = (physerv_ne - physerv_ne_lag12) / physerv_ne_lag12    
#   ) 
# testdataset <- testdataset %>%
#   mutate(
#     quantitygrowth = expgrowth - pricegrowth,
#     pricegrowth = pricegrowth*100,
#     quantitygrowth = quantitygrowth*100)
# 
# components <- c("pricegrowth", "quantitygrowth")
# # Prepare Data
# teststack <- prep_stacked_data(
#   data = testdataset,
#   contrib_cols = components,
#   display_names = c(
#   "Price",
#   "Quantity"
#   )
# )
# 
# 
# # Set Up Plot Specifications
# testplotspecs <- data.frame(
#   title = c(
#     "Healthcare Services Expenditure Decomposition",
#     "Healthcare Services Expenditure Decomposition since 2000"
#   ),
#   filename = c(
#     "hcserv_exp_decomp.png",
#     "hcserv_exp_decomp_2000.png"
#   ),
#   start_date = c(
#     NA, 
#     "2000-01-31"
#     ),
#   stringsAsFactors = FALSE
# )
# 
# # Graph
# for (i in 1:nrow(testplotspecs)) {
#   test <- plot_decomp(
#     data = teststack,
#     component_order = c(
#    "Price",
#     "Quantity"
#       ),
#     colors = c(
#      "Price" = "purple",
#      "Quantity" = "darkorange"
#       ),
#     title = testplotspecs$title[[i]],
#     start_date = if(!is.na(testplotspecs$start_date[i])) as.Date(testplotspecs$start_date[i]) else NULL
#     )
#   print(test)
#   # save_plot(test, testplotspecs$filename[[i]])
# }
# 
# 
# testdataset <- mutate(testdataset, total2 = expgrowth*100)


```








### Graph Decomposition of Healthcare Services Inflation
```{r}

# Prepare Data
g4stacked <- prep_stacked_data(
  data = g4data, 
  contrib_cols = paste0(components, "_contrib"),
  display_names = c(
  "Physician",
  "Dental", 
  "Paramedical", 
  "Hospital", 
  "Nursing Homes"
  )
)

# Set Up Plot Specifications
g4plotspecs <- data.frame(
  title = c(
    "Healthcare Services Inflation Decomposition",
    "Healthcare Services Inflation Decomposition since 2000"
  ),
  filename = c(
    "hcserv_infl_decomp.png",
    "hcserv_infl_decomp_2000.png"
  ),
  start_date = c(
    NA, 
    "2000-01-31"
    ),
  stringsAsFactors = FALSE
)

# Graph
for (i in 1:nrow(g4plotspecs)) {
  g4 <- plot_decomp(
    data = g4stacked,
    component_order = c(
      "Hospital",
      "Physician",
      "Paramedical",
      "Dental",
      "Nursing Homes"
      ),
    colors = c(
     "Hospital" = "purple",
      "Physician" = "darkblue",
     "Paramedical" = "darkorange",
     "Dental" = "forestgreen",
      "Nursing Homes" = "firebrick"
      ),
    title = g4plotspecs$title[[i]],
    start_date = if(!is.na(g4plotspecs$start_date[i])) as.Date(g4plotspecs$start_date[i]) else NULL
    )
  print(g4)
  save_plot(g4, g4plotspecs$filename[[i]])
}


```

### Clean Up G4 Values
```{r}
rm(g4data, g4plotspecs, g4stacked, g4, components, contrib_cols, display_names)
```


##################################################################################################################################################################################################################################

## (5) Core PCE Inflation Decomposition
```{r}
# Pull In Data
usnalist <- c("cxfem","cgxfem", "csxhem", "csdm", "jcxfem", "jcgxfem", "jcsxhem", "jcsdm")
rename_list <- c("core_ne", "goodxfe_ne", "servxeh_ne", "hcserv_ne", "core_pi", "goodxfe_pi", "servxeh_pi", "hcserv_pi")
g5data <- get_haver_data(usnalist, rename_list = rename_list)

#Find inflation, shares, and contributions
components <- c("goodxfe", "servxeh", "hcserv")
g5data <- g5data %>%
  calc_infl() %>%
  calc_shares(aggvar = "core") %>%
  calc_contrib() %>%
  calc_aggregate(aggvar = "core")



g5diffwcore <- g5data %>%
  mutate(year = format(date, "%Y")) %>%
  group_by(year) %>%
  summarize(
    hcserv_pi = mean(hcserv_pi),
    hcserv_expshare = mean(hcserv_expshare),
    core_pi = mean(core_pi))

write.dta(g5diffwcore, "C:/Users/g1sxh05/FRBC Dropbox/Shria Holla/KM_SH/Healthcare PCE/code/g5diffwcore.dta")

```

### Core PCE Inflation Decomposition
```{r}
# Prepare Data
g5stacked <- prep_stacked_data(
  data = g5data, 
  contrib_cols = paste0(components, "_contrib"),
  display_names = c(
  "Goods excl Food & Energy",
  "Services excl Energy & Healthcare", 
  "Healthcare Services"
  )
)

# Graph Full Date Range
g5 <- plot_decomp(
  data = g5stacked,

  component_order = c(
    "Healthcare Services",
    "Services excl Energy & Healthcare",
    "Goods excl Food & Energy"
    ),

  colors = c(
    "Healthcare Services" = "darkorange",
    "Services excl Energy & Healthcare" = "forestgreen",
    "Goods excl Food & Energy" = "darkblue"
  ),

  title = "Core PCE Inflation Decomposition"
)

print(g5)
save_plot(g5, "core_infl_decomp.png")
```


### Graph Share of Core PCE Inflation Contributed by Healthcare Services, Monthly
```{r}
# Prep Data
g5m <- g5data %>%
  mutate(hcserv_contrib_share = (hcserv_contrib/core_infl_est)*100)

g5msave <- select(g5m, date, core_infl_est, hcserv_contrib, hcserv_contrib_share)
write.csv(g5msave, file = "hcserv_infl_core_infl_share.csv", row.names = FALSE)


#Graph
g5m_share <- plot_line(
  data = g5m,
  y_cols = list(
    "Share" = "hcserv_contrib_share"
  ),
  colors = c(
    "Share" = "darkorange"
  ),
  title = "Share of Core PCE Inflation Contributed by Healthcare Services\nMonthly",
  y_label = "Percent (%)",
  show_legend = FALSE,
)

print(g5m_share)
save_plot(g5m_share, "hcserv_infl_share_core_infl.png")
```

### Graph Share of Core PCE Inflation Contributed by Healthcare Services, Annual
```{r}
#Annual
## SH26 have to make this automatic! hard coded the dates!
g5a <- g5data %>%
  mutate(year = format(date, "%Y"),
         hcserv_contrib_share = (hcserv_contrib/core_infl_est)*100) %>%
  group_by(year) %>%
  summarize(yrlyshare = mean(hcserv_contrib_share))

g5a_share <- plot_line_year(
  data = g5a,
  x_col = "year",
  y_cols = list(
    "Share" = "yrlyshare"
  ),
  colors = c(
    "Share" = "darkorange"
  ),
  title = "Share of Core PCE Inflation Contributed by Healthcare Services Inflation\nAnnual",
  y_label = "Percent (%)",
  show_legend = FALSE,
  x_breaks = c(seq(1960, 2024, by = 2), 2026),
  start_date = 1960
  
)

print(g5a_share)
save_plot(g5a_share, "hcserv_infl_share_core_infl_annual.png")
```

### Clean Up G5 Values
```{r}
rm(g5, g5data, g5a, g5m, g5stacked, g5m_share, g5msave, g5a_share, components, contrib_cols, display_names)
```



##################################################################################################################################################################################################################################
## (6) Overall PCE Inflation Decomposition

```{r}
# Pull In Data
usnalist <- c("cm","ctgm", "csm", "csdm", "jcm", "jctgm", "jcsm", "jcsdm")
rename_list <- c("ov_ne", "good_ne", "serv_ne", "hcserv_ne", "ov_pi", "good_pi", "serv_pi", "hcserv_pi")
g6data <- get_haver_data(usnalist, rename_list = rename_list)

#Find inflation, shares, and contributions
components <- c("good", "serv", "hcserv")
g6data <- g6data %>%
  calc_infl() %>%
  calc_shares(aggvar = "ov") %>%
  calc_contrib()

#Calculate Contribution of Services w/out Healthcare Services
g6data$servxh_contrib <- g6data$serv_contrib - g6data$hcserv_contrib

#Aggregate contributions
components <- c("good", "servxh", "hcserv")
g6data <- calc_aggregate(g6data, aggvar = "ov")

```


### Graph Overall PCE Decomposition
```{r}
# Prepare Data
g6stacked <- prep_stacked_data(
  data = g6data, 
  contrib_cols <- paste0(components, "_contrib"),
  display_names <- c(
  "Goods",
  "Services excl Healthcare", 
  "Healthcare Services"
  )
)

#Plot Specifications
g6plotspecs <- data.frame(
  title = c(
    "Overall PCE Inflation Decomposition",
    "Overall PCE Inflation Decomposition since 2000"
  ),
  start_date = c(
    NA, 
    "2000-01-31"
  ),
  filename = c(
    "ov_infl_decomp.png",
    "ov_infl_decomp_2000.png"
  )
)

#Graph
for (i in 1:nrow(g6plotspecs)) {
  g6 <- plot_decomp(
    data = g6stacked,
    component_order = c(
      "Healthcare Services",
      "Services excl Healthcare",
      "Goods"
    ),
    colors = c(
      "Healthcare Services" = "darkorange",
      "Services excl Healthcare" = "forestgreen",
      "Goods" = "darkblue"
    ),
    title = g6plotspecs$title[[i]],
    start_date = if(!is.na(g6plotspecs$start_date[i])) as.Date(g6plotspecs$start_date[i]) else NULL 
  )
  print(g6)
  save_plot(g6, g6plotspecs$filename[[i]])
}

```

### Clean Up G6 Values
```{r}
rm(g6data, g6plotspecs, g6stacked, g6, components, contrib_cols, display_names)
```



##################################################################################################################################################################################################################################
## (7) Graph Collection of Price Indexes

```{r}

# For PCE series
pce_series <- data.frame(
  code = c("jcsdm", "jcsdom", "jcsmpm", "jcsmdm", "jcsmom", "jcsmnm", "jcsmhpm", "jcsmhnm"),
  name = c("hcserv", "opserv", "physerv", "dentserv", "paraserv", "hnserv", "hosp", "nurs"),
  display = c("Healthcare Services", "Outpatient Services", "Physician Services", "Dental Services", "Paramedical Services", "Hospital and Nursing Home Services", "Hospital Services", "Nursing Home Services"),
  color = c("blue", "green", "darkblue", "forestgreen", "darkorange", "deeppink", "purple", "firebrick")
)


pce_results <- plot_price_indices(
  series_data = pce_series,
  index_type = "PCE"
)

# For PPI series
ppi_series <- data.frame(
  code = c("r6211", "r6216", "r6215", "r622", "r6231"),
  name = c("physerv", "hhcserv", "medlab", "hosp", "nurs"),
  display = c("Offices of Physicians", "Home Health Care Services", "Medical and Diagnostic Labs", "Hospitals", "Nursing Care Facilities"),
  color = c("darkblue", "darkorange", "darkorange", "purple", "firebrick")
)


ppi_results <- plot_price_indices(
  series_data = ppi_series,
  index_type = "PPI"  # Uses PPIR database
)


# For CPI series
cpi_series <- data.frame(
  code = c("umsptn", "umspsn"),
  name = c("dentserv", "othmpserv"),
  display = c("Dental Services", "Other Medical Professionals' Services"),
  color = c("forestgreen", "darkorange")
)

cpi_results <- plot_price_indices(
  series_data = cpi_series,
  index_type = "CPI"  # Uses CPIDATA database
)

```

#Clean Up G7 Values
```{r}
rm(cpi_results, cpi_series, pce_results, pce_series, ppi_results, ppi_series)
```


